---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(MotrpacHumanPreSuspensionAnalysis)
```

# MotrpacHumanPreSuspensionAnalysis

<!-- badges: start -->
<!-- badges: end -->

# Installation

More vignettes are currently being developed, as well as a pkgdown website to get users more familiar with the data. 

```{r, eval = FALSE}
devtools::install_github("MoTrPAC/MotrpacHumanPreSuspensionAnalysis",
                         build_vignettes = TRUE)
```

We recommend building the vignettes when installing. 

(use `vignette(package = "MotrpacHumanPreSuspensionAnalysis")`)


## Package Use and Structure

This package is the public release version of the results from the Molecular Transducers of Physical Activity Consortium. The primary aim of the package is to facilitate enabling reproducible downstream interpretation and access to analysis results. 

To protect participant privacy and comply with data-use governance policies, individual-level (subject-level) molecular or phenotypic data are not included in this package and are available only through formal data access requests to the MoTrPAC consortium. This package will instead include the n, mean, and SD for feature in each exercise group and tissue as well as the differential analysis/hypothesis testing summary statistics. It will further include various metadata about the feature to gene mapping or sample outliers.

## Getting help

For questions, bug reporting, and data requests for this package, please 
[submit a new issue](https://github.com/MoTrPAC/MotrpacHumanPreSuspensionAnalysis/issues){target="_blank"} 
and include as many details as possible. 


###  Omic modeling summary statistics (Differential Analysis)

See more documentation via `?load_differential_analysis`

Note that the public release of epigenetic files is through AWS CDN, and the file sizes are quite a bit larger than the other omes, so setting `epigen = TRUE` can be very slow!

```{r differential analysis}
differential_analysis = load_differential_analysis(
  selected_omes = "all",
  selected_tissues = "all",
  single_matrix = FALSE,
  epigen = FALSE,
  combine_with_featgene = FALSE,
  verbose = TRUE
)
names(differential_analysis)
names(differential_analysis[["blood"]])
```

By default,`load_differential_analysis` loads in the dataset in a nested list first by tissue, then by ome. Choose whichever tissues or omes you'd like via `selected_omes` or `selected_tissues`. You can find available tissues via `tissue_available_list()` or `ome_available_list()`. Or if you enter in a wrong mistaken tissue/ome, a warning or error will help.

If you would instead like to stack the matrixes more easily, use the `single_matrix` function, which basically unlists the list and sticks everything into a data.frame object.

```{r}
single_matrix = load_differential_analysis(single_matrix = TRUE)
colnames(single_matrix)
```
For a quick explanation of each of the columns, you can find this via `?load_differential_analysis`

Importantly, this loads the differential analysis for each of comparisons mentioned in the methods, including the comparison between the endurance or resistance group relative to time, fasting, biopsy, etc. matched controls, comparison between the endurance and resistance groups directly, and finally comparisons within group without a matched control. 

The majority of the analysis is done via exercise groups relative to the controls ("exercise_with_controls"). Make sure you filter to whichever category you prefer before continuing with analysis. 

```{r}
single_matrix %>% dplyr::pull(contrast_type) %>% unique()
```
If you'd like to display things in terms of the specific groups being compared instead, you can use the 'contrast_category' column. (EE-CON, RE-CON would be subsets of the 'exercise_with_controls' category from above, for example.)
```{r}
single_matrix %>% dplyr::pull(contrast_category) %>% unique()
```

The splicing data was processed in a separate analysis effort, but significant results (FDR < 0.05) are available in this R package as well. The full set is available on the motrpac data hub, but is not included here because of file size limitations. See "Exercise modulation of the alternative splicing landscape in human tissues" for more information. 

```{r}
names(SPLICING_DA)
head(SPLICING_DA$adipose$`AS-rMATS`, 3)
```


### Omic modeling summary statistics 

See more documentation via `?load_summary_stats`

```{r summary stats}
summary_stats = load_summary_stats(
  selected_omes = "all",
  selected_tissues = "all",
  single_matrix = FALSE,
  verbose = TRUE
)
names(summary_stats)
names(summary_stats[["blood"]])
```

By default, load_summary_stats() loads group- and timepoint-level summary statistics for normalized expression data in a nested list structure, organized identically to the differential-analysis datasets. The top level corresponds to tissues, and the second level corresponds to molecular assays or platforms. 

You may subset the data using selected_tissues and selected_omes. Available options can be queried via `tissue_available_list()` and `ome_available_list()`. If an invalid tissue or assay is supplied, informative warnings or errors are raised to guide correction.

If a stacked representation is preferred, set `single_matrix = TRUE`. This unlists the nested structure and returns a single `data.frame` with all selected tissues and assays.

Again - note that the sample level data is available for researchers upon request via the Motrpac Consortium. 

```{r}
single_matrix = load_summary_stats(single_matrix = TRUE)
colnames(single_matrix)
```

Although summary statistics are provided for all normalized expression features, only features with paired sample size n >= 3 were eligible for hypothesis testing in downstream differential analyses. As a result, some features—particularly those affected by missingness in certain assays—may appear in the summary-statistics tables but lack corresponding differential-analysis p-values.

For metabolomics assays, summary statistics are computed after filtering redundant metabolites. Details of this filtering procedure are described in the Methods section of the manuscript. 


## Other items


### Enrichment Results
```{r}
colnames(CAMERA_RESULTS)
```
Quick summary: CAMERA-PR is a method of enrichment that incorporates all features to generate a comparison of the test statistics between in-pathway vs out-of-pathway test statistics to see if the statistics within pathway are significant.

This file structure is more or less just an enrichment level match for the comparisons described in the single-matrix differential analysis results, where all tissues and assays are included in all the analysis.


### Feature to gene file: 
```{r}
head(HUMAN_FEATURE_TO_GENE)
```

The feature-to-gene map links each feature tested in differential analysis to a gene, using Ensembl version 105 (mapped to GENCODE 39) as the gene identifier source. Proteomics feature IDs (UniProt IDs) were mapped to gene symbols and Entrez IDs using UniProt’s mapping files. Epigenomics features were mapped to the nearest gene using the ChIPseeker::annotatePeak() function with Homo sapiens Ensembl release 105 gene annotations. Gene symbols, Entrez IDs, and Ensembl IDs were assigned to features using biomaRt version 2.58.2 (Bioconductor 3.18). This file links all of the features included in any ome/tissue in our analysis. Use this to see how some levels of omic analysis (e.g. ATAC, RNAseq) may link up in terms of ome names. 


## Acknowledgements 

MoTrPAC is supported by the National Institutes of Health (NIH) Common
Fund through cooperative agreements managed by the National Institute of Diabetes and Digestive and Kidney Diseases (NIDDK), National Institute of Arthritis and Musculoskeletal Diseases (NIAMS), and National Institute on Aging (NIA). 

Specifically, the MoTrPAC Study is supported by NIH grants U24OD026629 (Bioinformatics Center), U24DK112349, U24DK112342, U24DK112340, U24DK112341, U24DK112326, U24DK112331, U24DK112348 (Chemical Analysis Sites), U01AR071133, U01AR071130, U01AR071124, U01AR071128, U01AR071150, U01AR071160, U01AR071158 (Clinical Centers), U24AR071113 (Consortium Coordinating Center), U01AG055133, U01AG055137 and U01AG055135 (PASS/Animal Sites).

## Data Use Agreement 

Recipients and their Agents agree that in publications using **any** data from MoTrPAC public-use data sets they will acknowledge MoTrPAC as the source of data, including the version number of the data sets used, e.g.:

* Data used in the preparation of this article were obtained from the Molecular Transducers of Physical Activity Consortium (MoTrPAC) database, which is available for public access at [motrpac-data.org](motrpac-data.org). 
* Data used in the preparation of this article were obtained from the Molecular Transducers of Physical Activity Consortium (MoTrPAC) Pre-CovidSuspension Data release version 1.3.0.

