---
title: "MotrpacHumanPreSuspensionAnalysis: Differential Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MotrpacHumanPreSuspensionAnalysis: Differential Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MotrpacHumanPreSuspensionAnalysis)
```

# Overview

This vignette demonstrates how to use `load_differential_analysis()` to access differential analysis (DA) results generated by the MoTrPAC Human Pre-COVID sedentary adult study. The function provides a unified interface for loading curated summary statistics across multiple molecular assays (“omes”) and tissues, with consistent column structure and metadata harmonization.

The package is designed to distribute summary-level results only. Subject-level data are not included and must be requested separately through MoTrPAC data access procedures.


## Basic Usage

```{r load-da-all}
DA_list <- load_differential_analysis()
```

This loads all available tissues and omes using lazily loaded package data. Results are returned as a nested list, organized by tissue and assay.

To inspect a single result table:
```{r da-adipose-prot-str}
str(DA_list[["adipose"]][["prot-pr"]])
```

## Selecting Tissues and Omics Modalities

Users can restrict loading to specific tissues and omes to reduce memory usage and improve clarity.

```{r load-da-muscle}
DA_list <- load_differential_analysis(
  selected_tissues = c("muscle"),
  selected_omes = c("transcript-rna-seq", "prot-pr")
)
```

Internally, metabolomics platforms are handled with special care: if any metabolomics assay is requested, all metabolomics results are loaded to ensure consistent platform-level filtering.

Note: For metabolomics assays, multiple features may map to the same reference metabolite. To reduce redundancy:

The lowest coefficient-of-variation (CV) feature is selected.

Feature identifiers are replaced with standardized RefMet names.

P-values are recalculated within contrasts after filtering.

Users should consult the `METABOLOMICS_CVS` object if a metabolite of interest appears to be missing.

If you forget which syntax is used for omes & tissues:

```{r available-omes-tissues}
ome_available_list()
tissue_available_list()
```


## Epigenomics Data (ATAC-seq and MethylCap-seq)

Epigenomics results are not stored directly in the package due to file size constraints. When `epigen = TRUE`, results are downloaded from the MoTrPAC Google Cloud bucket using `gsutil` and cached locally.

```{r load-da-epigen, eval = FALSE}
DA_list <- load_differential_analysis(
  epigen = TRUE,
  repo_local_dir = "/path/to/local/cache"
)
```

*Important points to note*:

Epigenomics downloads can take 30 minutes or longer.

A local directory is required for temporary file storage.

Users must have working `gsutil` access and appropriate permissions.


## Splicing Data

A data object containing results from a differential alternative splicing
analysis. The analytical workflow used to generate these results—including
read alignment, isoform quantification, splicing event detection, and
statistical testing—is described in detail in: "Characterization of exercise-modulated
alternative splicing landscape in human skeletal muscle, adipose tissue, and blood in the MoTrPAC Study". 
Due to file size limitations, only significant (FDR<0.05) results are included as a data object.


```{r splicing-da-names}
names(SPLICING_DA)
```

## Including the results as a larger matrix instead

Since the differential abundance analysis largely all comes in one format, you can choose to display the results in one big matrix format instead, if you find that to be easier to work with.

```{r load-da-single-matrix}
DA_matrix <- load_differential_analysis(
  single_matrix = TRUE
)
```


## Feature-to-Gene Annotation

Setting `combine_with_featgene = TRUE` merges DA results with the
`HUMAN_FEATURE_TO_GENE` mapping table, adding gene-level annotations where applicable.

```{r load-da-annotated}
DA_list <- load_differential_analysis(
  combine_with_featgene = TRUE
)
```


This is particularly useful for transcriptomic and proteomic assays when integrating results across omes.


