---
title: "MotrpacHumanPreSuspensionAnalysis: Complete Package Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MotrpacHumanPreSuspensionAnalysis: Complete Package Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(MotrpacHumanPreSuspensionAnalysis)
```

---

# Introduction

`MotrpacHumanPreSuspensionAnalysis` is an R package that distributes
summary-level results from the MoTrPAC Human Pre-COVID sedentary adult
exercise study. It provides a unified interface for loading, exploring, and
analyzing differential abundance results across multiple molecular assays
("omes") and tissues, along with tools for gene set enrichment analysis,
fuzzy c-means clustering, and publication-quality visualization.

**Important:** This package distributes summary-level results only.
Subject-level data are not included and must be requested separately through
[MoTrPAC data access procedures](https://motrpac-data.org).

## Study design overview

The study examined acute exercise responses across three tissues (adipose,
blood, and muscle) in sedentary adults randomized to endurance exercise (EE),
resistance exercise (RE), or control (CON) groups. Samples were collected at
multiple timepoints relative to a single bout of exercise:

- Pre-exercise (baseline)
- During exercise (20 min, 40 min)
- Post-exercise (10 min, 15/30/45 min, 3.5/4 hr, 24 hr)

Molecular profiling spanned seven assay types:

- Transcriptomics (RNA-seq)
- Global proteomics (prot-pr)
- Phosphoproteomics (prot-ph)
- Olink proteomics (prot-ol)
- Metabolomics (14 targeted and untargeted platforms)
- ATAC-seq (epigen-atac-seq)
- MethylCap-seq (epigen-methylcap-seq)

---

# 1. Available Tissues and Omes

Before loading any data, you can check which tissues and omes are available.

```{r}
tissue_available_list()
```

```{r}
ome_available_list()
```

To see only the metabolomics platforms:

```{r}
metab_only_list()
```

---

# 2. Data Objects

The package ships with a variety of lazily loaded data objects. These objects
are the backbone of most functions in the package.

## 2.1 Differential Analysis Results

The core DA results are stored as internal data objects (e.g.,
`MUSCLE_PROT_PR_DA`, `BLOOD_TRNSCRPT_DA`). These should **not** be accessed
directly. Instead, use the `load_differential_analysis()` function described
below.

## 2.2 Summary Statistics

Group-level means and standard deviations for normalized expression data are
stored as `*_SUM_STATS` objects (e.g., `MUSCLE_PROT_PR_SUM_STATS`). Use
`load_summary_stats()` to access these.

## 2.3 Contrast Converter

The `CONTRAST_CONVERTER` table maps all **33 contrasts** used in differential
analysis to their short names, types, and categories. Understanding the
contrast structure is essential for correctly interpreting results.

For a graphical overview of the study design and timepoints, see the
[MoTrPAC Acute Exercise Study Design](https://motrpac-data.org/project-overview#acute-exercise).

```{r}
head(CONTRAST_CONVERTER)
```

### Contrast types

The 33 contrasts are organized into five types, each addressing a different
biological question:

- **`exercise_with_controls`** (10 contrasts) — The primary analysis.
  Compares each exercise group (Endurance or Resistance) to time-matched
  controls using a delta-delta design: (exercise post - exercise pre) -
  (control post - control pre). This removes baseline differences and
  isolates the exercise-specific response.

- **`exercise_no_controls`** (10 contrasts) — Within-group change from
  baseline for each exercise group (post - pre), without adjusting for
  control group changes. Useful when control timepoints are unavailable
  (e.g., during-exercise samples at 20 and 40 min).

- **`Endur_vs_Resist`** (4 contrasts) — Direct comparison of endurance vs.
  resistance exercise at each post-exercise timepoint (delta-delta design).
  Identifies modality-specific responses.

- **`baseline`** (3 contrasts) — Pre-exercise differences between groups
  (Endurance vs. Resistance, Endurance vs. Control, Resistance vs. Control).
  Verifies randomization and identifies pre-existing group differences.

- **`control_only`** (6 contrasts) — Control group changes over time relative
  to their own pre-exercise baseline. Captures effects of fasting, biopsy,
  or circadian variation unrelated to exercise.

### Available contrasts by type

```{r, fig.width = 6, fig.height = 3}
# Number of contrasts per type
contrast_counts <- as.data.frame(table(CONTRAST_CONVERTER$contrast_type))
colnames(contrast_counts) <- c("contrast_type", "n")

ggplot2::ggplot(contrast_counts,
                ggplot2::aes(x = reorder(contrast_type, n), y = n)) +
  ggplot2::geom_col(fill = "#3C5488") +
  ggplot2::geom_text(ggplot2::aes(label = n), hjust = -0.3, size = 3.5) +
  ggplot2::coord_flip() +
  ggplot2::labs(x = NULL, y = "Number of contrasts",
                title = "Contrasts per type (33 total)") +
  ggplot2::theme_minimal() +
  ggplot2::ylim(0, max(contrast_counts$n) + 1)
```

```{r, results = "asis"}
# Show the short contrast names grouped by type
for (ct in unique(CONTRAST_CONVERTER$contrast_type)) {
  cat("**", ct, "**\n\n", sep = "")
  shorts <- CONTRAST_CONVERTER$contrast_short[CONTRAST_CONVERTER$contrast_type == ct]
  for (s in shorts) cat("-", as.character(s), "\n")
  cat("\n")
}
```

Most downstream analyses focus on `exercise_with_controls`. Make sure to
filter by `contrast_type` before interpreting results:

```{r, eval = FALSE}
DA <- load_differential_analysis(single_matrix = TRUE)

# Keep only the primary exercise-vs-control contrasts
DA_primary <- DA[DA$contrast_type == "exercise_with_controls", ]
```

## 2.4 Feature-to-Gene Mapping

`HUMAN_FEATURE_TO_GENE` is a large lookup table (~1.96M rows) that maps
feature IDs to gene symbols, Entrez IDs, Ensembl IDs, UniProt IDs, RefMet
names, KEGG IDs, and flanking sequences (for phosphoproteomics).

```{r, eval = FALSE}
str(HUMAN_FEATURE_TO_GENE[1:5, ])
```
which should print:

```
Classes ‘data.table’ and 'data.frame':	5 obs. of  11 variables:
 $ assay               : Factor w/ 7 levels "epigen-atac-seq",..: 1 1 1 1 1
 $ feature_id          : Factor w/ 1958568 levels "(E,E)-Trichostachine",..: 793 794 795 796 797
 $ entrez_gene         : Factor w/ 23406 levels "1","10","100",..: 8521 8521 4916 8521 8521
 $ gene_symbol         : Factor w/ 32911 levels "5_8S_rRNA","5S_rRNA",..: 27108 27108 22521 27108 27108
 $ ensembl_gene        : Factor w/ 50007 levels "ENSG00000000003",..: 4704 4704 10604 4704 4704
 $ custom_annotation   : Factor w/ 10 levels "3' UTR","5' UTR",..: 6 5 6 6 6
 $ relationship_to_gene: num  0 0 0 0 0
 $ uniprot             : Factor w/ 10736 levels "A0A075B6H7","A0A075B6H8",..: NA NA NA NA NA
 $ refmet_name         : Factor w/ 1513 levels "(E,E)-Trichostachine",..: NA NA NA NA NA
 $ kegg_id             : Factor w/ 392 levels "C00002","C00003",..: NA NA NA NA NA
 $ flanking_sequence   : Factor w/ 30682 levels "------MsDEEVEQV",..: NA NA NA NA NA
 - attr(*, ".internal.selfref")=<externalptr> 
 - attr(*, "sorted")= chr [1:2] "assay" "feature_id"
 ```
 
## 2.5 Metabolomics CVs

`METABOLOMICS_CVS` contains coefficients of variation for metabolomics
features, used internally to select the lowest-CV feature when multiple
features map to the same RefMet metabolite.

```{r}
head(METABOLOMICS_CVS)
```

## 2.6 Molecular Signatures

`MOLECULAR_SIGNATURES` is a named list of gene set databases used for
enrichment analysis. Available databases include:

```{r}
names(MOLECULAR_SIGNATURES)
```

These include canonical pathway databases (BIOCARTA, KEGG_MEDICUS, PID,
REACTOME, WikiPathways), Gene Ontology collections (GOBP, GOCC, GOMF),
MitoCarta3.0, PhosphoSitePlus kinase substrates (PSP), RefMet chemical
subclasses (REFMET), PTMSigDB, and CellMarker.

## 2.7 SET_TO_ID

`SET_TO_ID` maps each molecular signature to a unique numeric ID and
shortened name for use in figures and enrichment analyses.

```{r}
head(SET_TO_ID)
```

## 2.8 Pre-computed Enrichment and Clustering Results

The package includes pre-computed results for convenience:

- `CAMERA_RESULTS` — CAMERA-PR enrichment results across all tissues and omes
- `FCM_CLUSTERS` — Fuzzy c-means clustering assignments
- `FCM_CAMERA` — Cluster-level CAMERA enrichment results
- `FCM_ORA` — Cluster-level over-representation analysis results

## 2.9 Other Data Objects

- `COVARIATES_FILE` — covariates used in differential modeling
- `OUTLIERS` — flagged outlier samples from QC review
- `UTORONTO_TFs` — curated human transcription factor list (Lambert et al.)
- `OME_TISSUE_CODE` — tissue codes for each ome
- `SPLICING_DA` — significant differential alternative splicing results
  (FDR < 0.05)

### Color and Annotation Objects

For consistent figure styling, the package provides named color vectors:

- `HUMAN_TISSUE_COLORS` — tissue colors
- `HUMAN_EXERCISE_GROUP_COLORS` — exercise group colors (EE, RE, CON)
- `HUMAN_ACUTE_TIMEPOINT_COLORS` — timepoint phase colors
- `HUMAN_OME_COLORS` — molecular modality colors
- `HUMAN_SEX_COLORS` — sex-based colors
- `HUMAN_TISSUE_ABBR` — tissue abbreviations

---

# 3. Loading Differential Analysis Results

The primary entry point for accessing DA results is
`load_differential_analysis()`.

## 3.1 Load all results

```{r}
DA_list <- load_differential_analysis()
```

This returns a nested list organized by tissue (top level) and ome (second
level):

```{r}
names(DA_list)
lapply(DA_list, names)
```

Each element is a `data.table` with standardized columns including `tissue`,
`assay`, `contrast`, `contrast_type`, `feature_id`, `logFC`, `z.std`,
`p_value`, `adj_p_value`, and more.

```{r}
str(DA_list[["muscle"]][["prot-pr"]])
```

## 3.2 Filter by tissue and ome

```{r}
DA_muscle_rna <- load_differential_analysis(
  selected_tissues = "muscle",
  selected_omes = "transcript-rna-seq"
)
```

## 3.3 Return a single matrix

For analyses that prefer a flat data frame:

```{r}
DA_matrix <- load_differential_analysis(single_matrix = TRUE)
dim(DA_matrix)
```

## 3.4 Include gene annotations

Merge DA results with `HUMAN_FEATURE_TO_GENE` in one step:

```{r}
DA_annotated <- load_differential_analysis(
  selected_tissues = "muscle",
  selected_omes = "prot-pr",
  combine_with_featgene = TRUE
)
colnames(DA_annotated[["muscle"]][["prot-pr"]])
```

## 3.5 Epigenomics data

Epigenomics results are not bundled with the package due to file size.
Setting `epigen = TRUE` downloads them from AWS:

```{r, eval = FALSE}
DA_list <- load_differential_analysis(epigen = TRUE)
```

This requires network access and can take 30+ minutes for the initial
download.

## 3.6 Metabolomics handling

When any metabolomics platform is selected, all metabolomics data are loaded
to enable consistent platform-level filtering. Redundant features mapping to
the same RefMet metabolite are resolved by selecting the lowest-CV feature,
and p-values are recalculated within contrasts. If a metabolite of interest
appears missing, consult `METABOLOMICS_CVS`.

## 3.7 Splicing data

Differential alternative splicing results (FDR < 0.05 only) are available
directly:

```{r}
names(SPLICING_DA)
```

---

# 4. Loading Summary Statistics

`load_summary_stats()` loads group-level means and standard deviations for
normalized expression data.

```{r, eval = FALSE}
## Load all summary statistics
sum_stats <- load_summary_stats()

## Load metabolomics only
metab_stats <- load_summary_stats(selected_omes = "metab")

## Load as a single combined table
adipose_rna <- load_summary_stats(
  selected_tissues = "adipose",
  selected_omes = "transcript-rna-seq",
  single_matrix = TRUE
)
```

The interface mirrors `load_differential_analysis()` with `selected_tissues`,
`selected_omes`, and `single_matrix` parameters.

---

# 5. Visualization

The package provides several plotting functions for both individual features
and pathway-level results.

## 5.1 Single Feature Plots

`plot_single_feature()` creates time-course line plots for a given feature,
gene, or metabolite across tissues and omes. Points are highlighted when the
adjusted p-value falls below a threshold.

```{r, fig.width = 5, fig.height = 3}
# Returns a ggplot object that displays inline.
# Use the optional output_file parameter to save to PDF.
plot_single_feature(
  feature = "TFEB",
  selected_tissues = "muscle",
  selected_omes = "prot-pr",
  p_level = 0.05,
  color_time_labels = TRUE,
  include_legend = FALSE
)
```

Key parameters:

- `feature` — a feature_id, gene_symbol, or refmet_name (case-insensitive)
- `p_level` — adjusted p-value threshold; significant points are filled black
- `color_time_labels` — replaces x-axis text with colored tiles (single-panel
  only)
- `scale_factor` — scales all plot elements proportionally for
  posters/presentations
- `output_file` — optional path to save (recommended single-panel dimensions:
  H=1.54in, W=1.225in at scale_factor=1)

## 5.2 Feature Heatmaps

`plot_feature_heatmap()` generates a ComplexHeatmap for a set of features or
a molecular signature pathway, showing z-scores across contrasts.

```{r, eval = FALSE}
# Note: filename is required — the heatmap is saved directly to PDF.
# Heatmap from a specific pathway (using set_id from SET_TO_ID)
plot_feature_heatmap(
  set_id = "11725",
  selected_tissue = "muscle",
  selected_ome = "prot-ph",
  contrast_type = "exercise_with_controls",
  filename = "feature_heatmap_example.pdf"
)

# Heatmap from a custom list of features
plot_feature_heatmap(
  feature_ids = c("PPARGC1A", "TFEB", "VEGFA"),
  selected_tissue = "muscle",
  selected_ome = "transcript-rna-seq",
  filename = "custom_feature_heatmap.pdf"
)
```

Key parameters:

- `feature_ids` — character vector of feature IDs for the heatmap rows
- `set_id` — alternatively, a pathway ID from `SET_TO_ID` (overrides
  `feature_ids`)
- `selected_tissue` — one or more tissues (multi-tissue supported)
- `selected_ome` — exactly one ome
- `contrast_type` — which contrasts to display
- `post_min` / `post_hr` — specify exact timepoints (15, 30, or 45 min;
  3.5 or 4 hr)
- `full_modality_names` — display "Endurance Exercise" instead of "EE"
- `max_size` — limit the number of features displayed from a pathway

## 5.3 Enrichment Heatmaps

`plot_enrich_heatmap()` creates bubble heatmaps from CAMERA-PR or PTM-SEA
results, showing enrichment z-scores with significance bubbles.

```{r, eval = FALSE}
# Note: filename is required — the heatmap is saved directly to PDF.
# Using pre-computed CAMERA results
plot_enrich_heatmap(
  x = CAMERA_RESULTS,
  selected_tissues = "muscle",
  selected_ome = "prot-pr",
  contrast_type = "exercise_with_controls",
  n_top = 10,
  filename = "enrichment_heatmap.pdf"
)

# Filter to specific pathways
plot_enrich_heatmap(
  x = CAMERA_RESULTS,
  set_ids = c("00269", "00930"),
  selected_tissues = c("adipose", "blood"),
  selected_ome = "transcript-rna-seq",
  filename = "selected_pathways.pdf"
)
```

## 5.4 Cluster Trajectory Plots

After running fuzzy c-means clustering, use `plot_cmeans()` to visualize
cluster trajectories as line plots colored by membership probability.

```{r, eval = FALSE}
# Note: filename is required — the plot is saved directly to PDF.
FCM <- run_cmeans(selected_tissues = "adipose")

plot_cmeans(
  FCM = FCM[["adipose"]],
  filename = "adipose_clusters.pdf",
  min_membership = 0.3,
  ncol = 4
)
```

Key parameters:

- `keep_clusters` — subset to specific cluster numbers
- `min_membership` — minimum membership probability for display
- `common_ylim` — use the same y-axis across all cluster panels
- `modality` — show "both", "Endur" only, or "Resist" only

## 5.5 Cluster Enrichment Heatmaps

`plot_cluster_enrichment()` generates bubble heatmaps for cluster-level
enrichment results (from `run_cluster_cameraPR()` or `run_cluster_ORA()`).

```{r, eval = FALSE}
# Note: filename is required — the heatmap is saved directly to PDF.
plot_cluster_enrichment(
  x = FCM_CAMERA,
  selected_ome = "prot-pr",
  selected_tissues = "muscle",
  filename = "cluster_enrichment.pdf"
)
```

---

# 6. Enrichment Analysis

The package provides wrappers for competitive gene set testing and
over-representation analysis.

## 6.1 CAMERA-PR (Competitive Gene Set Testing)

`run_cameraPR()` performs pre-ranked Correlation Adjusted MEan RAnk gene set
analysis on the DA z-statistics. It tests whether molecular signatures show
coordinated up- or down-regulation.

```{r, eval = FALSE}
# Run on muscle global proteomics with GOBP and Reactome databases
camera_res <- run_cameraPR(
  selected_omes = "prot-pr",
  selected_tissues = "muscle",
  database = c("GOBP", "REACTOME")
)
head(camera_res)
```

```{r, eval = FALSE}
# Run on all omes with all databases (default)
camera_all <- run_cameraPR()
```

Key parameters:

- `database` — one or more of: BIOCARTA, KEGG_MEDICUS, PID, REACTOME, WP,
  GOBP, GOCC, GOMF, MITOCARTA, PSP (prot-ph only), REFMET (metab only)
- `path_to_gmt` — optionally supply custom GMT files (overrides `database`)
- `min_size` — minimum set size for testing (default: 5)
- `overlap_cutoff` — minimum proportion of set genes in the dataset
  (default: 0.7)
- `DA_list` — optionally supply custom DA results

Pre-computed results are available as `CAMERA_RESULTS`.

## 6.2 Over-Representation Analysis (ORA)

`run_ORA()` performs a hypergeometric test for enrichment of an input set of
"interesting" features against a background.

```{r}
# Define a background from all gene sets
bg <- MOLECULAR_SIGNATURES
bg[c("PSP", "REFMET")] <- NULL
bg <- unique(unlist(bg))

# Use OXPHOS genes as the interesting input
input <- MOLECULAR_SIGNATURES$MITOCARTA[["MITOCARTA_OXPHOS"]]

res <- run_ORA(input = input, background = bg)
head(res[, c("set", "set_size_in_input", "p_value", "adj_p_value")])
```

---

# 7. Fuzzy C-Means Clustering

`run_cmeans()` performs fuzzy c-means (FCM) clustering on z-score matrices
from differential analysis, grouping features with similar temporal response
patterns.

```{r, eval = FALSE}
# Run with defaults (13 clusters for adipose/blood, 12 for muscle)
FCM <- run_cmeans()
names(FCM)  # one element per tissue
```

```{r, eval = FALSE}
# Explore optimal cluster numbers
FCM_explore <- run_cmeans(
  selected_tissues = "adipose",
  num_clusters_adipose = 3:14
)

# Run for a single modality
FCM_endur <- run_cmeans(
  selected_tissues = "muscle",
  modality = "Endur"
)
```

Features are scaled by their standard deviation (including pre-exercise
zeros) before clustering. Clusters are reordered by hierarchical clustering
of centroids so that similar trajectories appear adjacent.

Pre-computed clustering results are available as `FCM_CLUSTERS`.

## 7.1 Cluster-Level Enrichment with CAMERA-PR

`run_cluster_cameraPR()` tests whether molecular signatures localize to
specific cluster centroids using modified signed rank tests.

```{r, eval = FALSE}
FCM <- run_cmeans()
cluster_camera <- run_cluster_cameraPR(FCM = FCM)
head(cluster_camera)
```

Pre-computed results: `FCM_CAMERA`.

## 7.2 Cluster-Level Over-Representation Analysis

`run_cluster_ORA()` performs ORA within each cluster, identifying
over-represented molecular signatures among features with high membership
probabilities.

```{r, eval = FALSE}
FCM <- run_cmeans()
cluster_ora <- run_cluster_ORA(
  FCM = FCM,
  min_prob = 0.3
)
head(cluster_ora)
```

Pre-computed results: `FCM_ORA`.

---

# 8. Typical Workflow

Below is a typical analysis workflow combining the major package functions:

```{r, eval = FALSE}
# 1. Load differential analysis results
DA_list <- load_differential_analysis(
  selected_tissues = "muscle",
  selected_omes = c("transcript-rna-seq", "prot-pr"),
  combine_with_featgene = TRUE
)

# 2. Explore a specific gene
plot_single_feature(
  feature = "PPARGC1A",
  selected_tissues = "muscle",
  selected_omes = c("transcript-rna-seq", "prot-pr"),
  p_level = 0.05
)

# 3. Run enrichment analysis
camera_res <- run_cameraPR(
  selected_omes = c("transcript-rna-seq", "prot-pr"),
  selected_tissues = "muscle",
  database = c("GOBP", "REACTOME")
)

# 4. Visualize top enriched pathways
plot_enrich_heatmap(
  x = camera_res,
  selected_tissues = "muscle",
  selected_ome = "prot-pr",
  contrast_type = "exercise_with_controls",
  n_top = 10,
  filename = "muscle_prot_enrichment.pdf"
)

# 5. Perform clustering
FCM <- run_cmeans(selected_tissues = "muscle")

# 6. Visualize clusters
plot_cmeans(
  FCM = FCM[["muscle"]],
  filename = "muscle_clusters.pdf",
  min_membership = 0.3
)

# 7. Cluster-level enrichment
cluster_res <- run_cluster_cameraPR(
  FCM = FCM,
  selected_tissues = "muscle"
)

# 8. Visualize cluster enrichment
plot_cluster_enrichment(
  x = cluster_res,
  selected_ome = "prot-pr",
  selected_tissues = "muscle",
  filename = "muscle_cluster_enrichment.pdf"
)
```

---

# 9. Session Info

```{r}
sessionInfo()
```
