---
title: "Human Precovid Metabolomics QC"
author: "Johanna Fleischman, Gayatri Iyer, Christopher Jin, David Jimenez-Morales, Daniel Katz and David Amar"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    toc_depth: 4
    code_folding: hide
---

## Summary

This notebook outlines a comprehensive quality control (QC) and analysis workflow for metabolomics data from the MoTrPAC human pre-COVID study. The analysis involves various stages to assess data integrity, perform customize normalization, and identify trends across the different metabolites. Each section contributes to refining the dataset for downstream analysis, ensuring the robustness and reliability of the conclusions drawn from these data. 

```{r setup, include=FALSE, echo=FALSE}
set.seed(1)
# set the current session
session_packages = c("tidyr","stringr","dplyr","tibble","knitr","kableExtra",
                     "limma","metap", "ggplot2","ggcorrplot",
                     "impute", "jsonlite",
                     "patchwork",
                     "eulerr","UpSetR",
                     "plotly","crosstalk","pheatmap","lme4","tibble",
                     "lmerTest","Hmisc","broom.mixed","furrr","purrr",
                     "MotrpacBicQC", "conflicted", "MatrixGenerics",
                     "MotrpacHumanPreSuspension")

for (lib_name in session_packages){
  tryCatch({library(lib_name,character.only = T)}, error = function(e) {
    print(paste("Cannot load",lib_name,", please install"))
  })
}
knitr::opts_chunk$set(echo = TRUE)

conflicted::conflicts_prefer(base::union)
conflicted::conflicts_prefer(MatrixGenerics::colIQRs)
conflicted::conflicts_prefer(base::setdiff)
conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(base::intersect)
conflicted::conflicts_prefer(base::setequal)

sessionInfo()
```

## Loading configuration parameters

This code uses a configuration file to ensure local file structure and local machine settings can be used properly.

```{r include=FALSE,message=FALSE,warning=FALSE}
#' STEPS FOR USING THE MoTrPAC CONFIG FILE:
#' 
#' 1. Create a "configs" FOLDER in your home directory. The location of this folder will depend on your operating system. For example:
#' - Windows: `C:\Users\youruser\configs`
#' - Mac: `/Users/youruser/configs`
#' - Linux: `/home/youruser/configs`
#' 2. Create a copy of `config-precovid-analyses.json` available in this repo in that folder and CUSTOMIZE the location for some of the variables.
#' 

# Detect the operating system
os_name <- Sys.info()["sysname"]

# Change default arguments if the OS is Windows
this_config <- file.path("configs", "config-precovid-analyses.json")

if (os_name == "Windows") {
  config_path <- file.path(Sys.getenv("USERPROFILE"), this_config) 
}else{
  config_path <- file.path(Sys.getenv("HOME"), this_config)  
}

# config <- jsonlite::read_json(path = config_path)
config = jsonlite::fromJSON("~/config.json")
repo_local_dir = config$precovid_repo_path


# Load libraries from `motrpac_bic_norm_qc`

#' Steps for using the MoTrPAC config file:
#' 1. Open the configuration template file in the root folder of this repo called `example_config.json` available in `precovid-analyses/`
#' 2. Save a copy of this template to some other location in your computer. Many choose to place it in their home directory: Example: `~/motrpac_config.json`
#' 3. Customize the fields for your machine and save changes. Remove any "comments" as JSON does not permit comments
#' 4. This script expects the following elements in the config file:
##' a. "gsutil_user": this is the way your machine calls the gsutil
##' b. "motrpac_bic_norm_qc_repo_path": this is the absolute location of the motrpac-bic-norm-qc repo
##' c. "precovid_repo_path": this is the absolute location of the motrpac-precovid-analyses repo
##' d. "general_shared_functions_library": this is the relative path location for the share functions library within the precovid repo
#' 5. Add the path to this config file here 
# config <- rjson::fromJSON(file = "/Users/gayat/Documents/MoTrPAC/motrpac_precovid_config_GI.json")

# config$motrpac_bic_norm_qc_repo_path = paste0(config$gitdir, "motrpac-bic-norm-qc/")
# config$precovid_repo_path = paste0(config$gitdir, "precovid-analyses/")
# Save date in case output files requiring date version
date <- Sys.Date()
date <- gsub("-", "", date) #we never actually refer to the date anywhere so we can maybe get rid of this, keeping for now

# Load libraries
source(file.path(normalizePath(config$motrpac_bic_norm_qc_repo_path),"tools/unsupervised_normalization_functions.R"))
source(file.path(normalizePath(config$motrpac_bic_norm_qc_repo_path),"tools/gcp_functions.R")) 
source(file.path(normalizePath(config$motrpac_bic_norm_qc_repo_path),"tools/qc_report_aux_functions.R"))
source(file.path(normalizePath(config$motrpac_bic_norm_qc_repo_path),"tools/config_session.R"))
source(file.path(normalizePath(config$motrpac_bic_norm_qc_repo_path),"tools/association_analysis_methods.R"))
source(file.path(normalizePath(config$motrpac_bic_norm_qc_repo_path),"tools/MetabolomicsLibrary.R"))
source(file.path(normalizePath(config$motrpac_bic_norm_qc_repo_path),"tools/metabolomics_data_parsing_functions.R"))

# Load relevant libraries in this repo
# source(file.path(normalizePath(config$precovid_repo_path), "library/library-general.R"))
source(file.path(normalizePath(config$precovid_repo_path), "library/JF-shared-functions.R"))
source(file.path(normalizePath(config$precovid_repo_path), "library/human_colors.R"))

tmp <- file.path(normalizePath(config$precovid_repo_path), "data", "tmp", "metabolomics_qc_norm")
if (!dir.exists(tmp)){
  dir.create(tmp)
}else{
  print("tmp exists")
}

gsutil_cmd <- config$gsutil_user
# gsutil_cmd = "gsutil"
```

In this document we analyze the Pre COVID metabolomics human data of MoTrPAC. The analysis flow is as follows:

1. Load the data from the google cloud (gs://motrpac-data-hub/)
2. Preprocess each dataset using our default normalization pipeline (details below)
3. Examine the mean/median intensities of the real and control samples
4. Perform PCA analysis for correlation with metadata variables
5. Flag and remove potential outliers
6. Save data for differential and integrated analysis

Note that metabolomics studies usually use quality control samples throughout the project for assessing the batches, and internal standards (spike-in metabolites) for normalization. 
We assume here that such normalization was done for each loaded dataset. 
Downstream normalization may be used to either deal with the heteroscedasticity of the metabolite-level data (e.g., Pareto normalization) or to account for sample-to-sample variation (MSTUS, Quantile, Median, etc.). 

First, specify pre-defined QC parameters

```{r qc metrics}

# Specify parameters for imputation process
imputation_alpha = 0.2 # changed to 0.2 on Oct 2020
max_allowed_NA_rate = 0.2 # changed to 0.2 on July 2023 to match PASS1B landscape

# Specify how many PCs to examine
num_pcs = 5
num_pcs_for_outlier_analysis = 3

# Specify the significance level for association analysis
# (e.g., between a PC and a clinical variable)
p_thr = 0.0001
OUTLIER_IQR_THR = 5

# Identify phenotypic metadata covariates to include in .csv output for
# integration/clustering analysis
# These variables will also be coerced into the sample_meta dataframe 
# even if the values are non-distinct for all parent samples of each daughter
required_meta_vars <- c("codedsiteid","visitcode","pid","BID","Timepoint","randomGroupCode",
                        "htcmavg_hwwt","wtkg_pcaa","calculatedAge","Sex")
```


## Load Data 

### Load metabolomics feature data and phenotype information 

*Skip this step* if you have already executed it 
(only run it if you need to update the data stored on your local machine). 

```{r load data from gcs, eval=F}
print(paste0("Data will be placed in ", tmp))
question1 <- readline("Would you like to redownload data from GCP? (Y/N)")

# Bucket structure is: tissue/platform/results/<data files>. For GET, this path should
# also have a qa_qc directory with the qc_metrics and sample_metadata files.
targeted_buckets = c("gs://motrpac-data-hub/human-precovid/results/metabolomics-targeted/")
untargeted_buckets = c("gs://motrpac-data-hub/human-precovid/results/metabolomics-untargeted/")
clinical_bucket = c(" gs://motrpac-data-hub/human-precovid/phenotype/human-precovid-sed-adu/raw/data_sets/")
pheno_bucket = "gs://motrpac-data-hub/human-eqc/adult/" #Human eqc data and randomization information

if(regexpr(question1, 'y', ignore.case = TRUE) == 1){
  for(targeted_bucket in targeted_buckets){
    rem_prev = targeted_bucket == targeted_buckets[1] # TRUE for first iteration of for loop, seems unnecessary
    obj = DownloadBucketLocal(targeted_bucket,tmp,gsutil_cmd)
  } 
  for(untargeted_bucket in untargeted_buckets){
    obj = DownloadBucketLocal(untargeted_bucket,tmp,gsutil_cmd)
  } 
  pheno_obj=DownloadBucketLocal(pheno_bucket,tmp,gsutil_cmd)
  clinical_obj = DownloadBucketLocal(clinical_bucket,tmp,gsutil_cmd)
} else {
  break
} 
```

### Load the parsed meta-data from local copy

This step is required for all analyses presented here.
In this block, data is parsed and stored in two objects used in downstream analyses:

1. metabolomics_parsed_datasets, which contains the peak/feature data and sample meta data parsed by tissue and platform

```{r load from local copy,message=FALSE,warning=FALSE,results="hide"}
# Create function input object of full pathnames to files, and local path
# This is separated by phenotype files and metabolomics data to prevent
# potential downstream errors

### GI update: Added additional 'gsub' statement to deal with difference between
### Windows and Mac in how they define file path ###
metab_ind = grepl("tmp/metabolomics",list.files(tmp, full.names = T, recursive = T))
metab_obj = list(downloaded_files = list.files(tmp, full.names = T, recursive = T)[metab_ind], 
                 local_path = tmp)
metab_obj$downloaded_files <- gsub("\\", "/", metab_obj$downloaded_files, fixed=T)
metab_obj$local_path <- gsub("\\", "/", metab_obj$local_path, fixed=T)
pheno_obj = list(downloaded_files = list.files(tmp, full.names = T, recursive = T)[!metab_ind], 
                 local_path = tmp)
pheno_obj$downloaded_files <- gsub("\\", "/", pheno_obj$downloaded_files, fixed=T)
pheno_obj$downloaded_files <- pheno_obj$downloaded_files[grepl("/tmp/adult", pheno_obj$downloaded_files)]
pheno_obj$local_path <- gsub("\\", "/", pheno_obj$local_path, fixed=T)

### End GI update ###

# Parse data by tissue and platform, store in nested list, separate from failed buckets (missing data or metadata, etc.) 
metab_datasets = read_metabolomics_datasets_from_download_obj(metab_obj,T)
metabolomics_parsed_datasets = metab_datasets$metabolomics_parsed_datasets
```

2. Phenotypic data from the `MotrpacHumanPreSuspension` package

```{r load pheno data,message=FALSE,warning=FALSE,results="hide"}
# Load phenotype data and dictionary for human adult EQC for precovid
pheno_data <- MotrpacHumanPreSuspension::load_pheno(
  repo_local_dir = tmp, 
  gsutil = config$gsutil_user, 
  load_acute_only = FALSE)

```

## Check for dataset issues

This block checks for errors in each of the parsed (tissue-platform) datasets in metabolomics_parsed_datasets.

This function will issue a warning or error if the following conditions are not met: 

1. The metabolite names are unique, 
2. All samples appear in meta data file, 
3. The annotation data and meta data have the same number of rows, 
4. All metabolites have annotation information,
5. All metabolites are named (no names are NA)

```{r check issues}
print(paste("Checking for issues in the datasets, we have",length(names(metabolomics_parsed_datasets)),"datasets"))
datasets_with_errors = c()

for(currname in names(metabolomics_parsed_datasets)){
  d = metabolomics_parsed_datasets[[currname]]
  
  # Fix site for Duke
  if(currname %in% c("t02-plasma,metab-t-conv,named", "t02-plasma,metab-t-imm-crt,named")){
    cat("- Fixing 'site' for ", currname, " (adding 'duke')\n")
    metabolomics_parsed_datasets[[currname]]$sample_meta$site <- "duke"
  }
  
  # The check_metabolomics_dataset_issues function takes the parsed data list item,
  # and the column index for desired row names (metabolite names). 
  # When rownames_ind = NULL the function will take the row names
  
  errs = check_metabolomics_dataset_issues(d, rownames_ind = 1)
  if(length(errs) > 0){
    print("####################################")
    print(currname)
    sapply(errs,print)
    datasets_with_errors = c(datasets_with_errors,currname)
  }
}
print(paste("Errors found in",length(datasets_with_errors),"datasets"))
```

## Data filtering and normalization

We go over each dataset and perform the following steps:

1. Average rows that have the same metabolite ID
2. Confirm that the metabolite annotation data and the sample intensity/concentration data have the same metabolite order
3. Merge the named and unnamed subparts of the untargeted datasets
4. Convert negative and zero values to missing
5. Remove features with $>20\%$ missing values
6. Features with $<20\%$ missing values are imputed using KNN
7. Data is log2-transformed 
8. Data is median-MAD normalized to account for sample-to-sample variation (untargeted only)

### Average rows with same metabolite ID and order annotation data and sample data

Data sets are checked to observe whether there are metabolites with the same name (duplicates). If that would be the case (which is not allowed within the same dataset), then the notebook stops. Then If enable, the average would be calculated. 

```{r data filtering,eval=TRUE,message=FALSE,warning=FALSE}

# Function averages the value of multiple metabolites with same name (as rows) in data frame
# Accepts data frame with first column as metabolite name and returns the data frame with merged values 
merge_dup_metabs <- function(temp_data) {
  merged_data = aggregate(temp_data[,-1], list(temp_data[,1]), mean, na.rm = T)  
  rownames(merged_data) = as.character(merged_data[,1])
  merged_data = merged_data[,-1]
  return(merged_data)
}

# Function returns annotation data in the same order as sample data (necessary if there is merging)
# Note that if a metabolite with the same name as different annotation data, one of these annotations
# will be lost in the averaging process
reconstruct_anno <- function(metabolites, anno) {
  if(all(metabolites %in% rownames(anno))){
    return(anno[metabolites,])
  }
  m = c()
  for(metab in metabolites) {
    ind = which(anno[,1] == metab)
    if(length(ind) < 1) {
      print(paste0("ERROR in reconstruct_anno: metabolite",
                   metab,"does not appear in the first column"))
      return(NULL)
    }
    ind = ind[1]
    m = rbind(m, anno[ind,])
    rownames(m)[nrow(m)] = metab
  }
  return(m)
}

# Merge metabolites (as an average) if they exist (multiple of the same name, as rows)
for(currname in names(metabolomics_parsed_datasets)) {
  message("PROCESSING: ", currname)
  d = metabolomics_parsed_datasets[[currname]]
  
  # Skip to next data set if there is only 1 metabolite in data set
  if(nrow(d$sample_data) < 2){
    message("\tSkipping dataset with only 1 metabolite\n\n")
    next
  } 
  new_control_data = NULL
  
  # If multiples of the same metabolite name exist in data set, stop
  if(any(table(d$sample_data[[1]]) > 1)) {
    stop("THERE ARE DUPLICATIONS IN THIS DATASET")
  }else {
    # ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
    # FILTERING STEP: 
    # Keep samples with a randomGroupCode VALUE available
    # ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
    valid_samples <- colnames(d$sample_data)[colnames(d$sample_data) %in% pheno_data$pheno_data$vialLabel[!is.na(pheno_data$pheno_data$randomGroupCode)]]
    new_sample_data = d$sample_data[,-1]
    rownames(new_sample_data) = d$sample_data[,1]
    new_sample_data = new_sample_data[,valid_samples]
    
    if(!is.null(dim(d$control_data)) && ncol(d$control_data) > 1) {
      new_control_data = d$control_data[,-1]
      rownames(new_control_data) = d$control_data[,1]
    }else{
      stop("REVIEW THE CODE! CONTROL DATA IS NOT IN THE EXPECTED FORMAT")
    }
  }
  
  # Replace sample data and control data with new, merged files 
  message("\t- Replaced data with reformatted data")
  d$sample_data = new_sample_data
  d$control_data = new_control_data
  
  # If the current data set was not flagged as having errors, order annotation  
  # and add an "is_named" column to indicate whether or not features are named
  if(!currname %in% datasets_with_errors){
    d$row_annot = reconstruct_anno(rownames(d$sample_data), d$row_annot)
    d$row_annot$is_named = ifelse(grepl(",named", currname), TRUE, FALSE)
  }
  
  # Replace working data set with new, merged data set
  metabolomics_parsed_datasets[[currname]] = d
}
gc()

# Re-check for data set issues
for(currname in names(metabolomics_parsed_datasets)){
  d = metabolomics_parsed_datasets[[currname]]
  if(nrow(d$sample_data) < 2) {next}
  errs = check_metabolomics_dataset_issues(d, NULL)
  if(length(errs) > 0) {
    print("####################################")
    print(currname)
    sapply(errs,print)
  }
}
print(paste(length(datasets_with_errors),"errors found in all datasets"))
```

### Merge the named and unnamed subparts of the untargeted datasets and normalize data

Here, we create a new list of objects, metabolomics_processed_datasets, in which named and unnamed datasets for tissue-platform combinations are merged. In these objects, there will be an entry called normalized_data, which is a list with different versions of the data:

* "raw" is the data received from CAS sites, with negative data values converted to missing (NA)
* "raw_imputed" data is the raw data with missing values imputed. Features with $>20\%$ missing values are removed. Features with $<20\%$ missing values are imputed using KNN
* "log2" data is the log2-transformed version of the "raw_imputed" data
* "log2_median_mad" data is the log2 data with median-MAD normalization applied to account for sample-to-sample variation. This normalization is applied on the untargeted datasets only

Note that for all log2 calculations for both raw mean intensities per sample and for datasets, 1 is added to each value before log-normalization. This allows metabolite values that fall between 0 and 1 to have a positive log2 value. Because raw_intensity per sample are calculated before cleaning and imputation (`raw_intensity_pre`) and post cleaning and imputation (`raw_intensity_post`) the log2 values on the `_pre` dataset may have negative values, as negative data has not been floored. 

This script also stores NA data (num_NAs in annotation data for each feature, and num_NAs in the sample meta data for each sample). The feature and sample IDs of these NA values are stored by feature and by sample in a separate list called na_data for datasets that have NA features.

**Here, if samples have missing samples IDs in the primary phenotype dataframe, we search for them in the parent-daughter data by BID. The print output indicates how many sample IDs were missing, and how many of these missing samples were matched to phenotype data by BID.**

```{r data merging,eval=T,message=FALSE}
# Initialize a object for storing new processed data sets
metabolomics_processed_datasets = c()

# Store the percent of features that were removed
percent_features_removed = c()
percent_remaining_values_imputed = c()

# Initialize a logical vector which indicates named or unnamed data sets to skip
# This is useful for skipping named or unnamed components of a data set that was already inspected 
metabolomics_raw_datasets_skip = rep(F,length(metabolomics_parsed_datasets))
names(metabolomics_raw_datasets_skip) = names(metabolomics_parsed_datasets)

i = 1
for(currname in names(metabolomics_parsed_datasets)) {
  if(metabolomics_raw_datasets_skip[currname]){
    message("\tSkipping dataset ", currname, "\n\n\n")
    next
  }
  
  print(paste0("STARTING ",currname," ------------------------------"))
  
  # Parse data set name by tissue and platform
  arr = strsplit(currname,split=",")[[1]]
  
  # Create both a named and unnamed string
  arr = arr[-length(arr)]
  name1 = paste(paste(arr,collapse=","),"named",sep=",")
  name2 = paste(paste(arr,collapse=","),"unnamed",sep=",")
  
  # If dataset is untargeted, named and unnamed are merged, and dataset is updated
  # If the merge fails, dataset is skipped and instead the error is printed
  if(name2 %in% names(metabolomics_parsed_datasets)){
    print("-- Merging named and unnamed data sets")
    d1 = metabolomics_parsed_datasets[[name1]]
    d2 = metabolomics_parsed_datasets[[name2]]
    new_processed_dataset = NULL
    tryCatch({
      new_processed_dataset = merge_named_and_unnamed_metabolomics_datasets(d1,d2,strict = F)
    }, error = function(e) {}) # can add error handling here
    
    # Add these data sets to skip list, so that they are not repeated
    metabolomics_raw_datasets_skip[name1] = TRUE
    metabolomics_raw_datasets_skip[name2] = TRUE
    
    # new_processed_dataset will only stay NULL as originally initialized if the merge function did not work
    if(is.null(new_processed_dataset)){
      print(paste("Skipping dataset:", currname,
                  "because merging the named and unnamed subsets failed"))
      stop(paste(currname, "Merging named and unamed failed. Please review the code"))
    }
  } else {
    metabolomics_raw_datasets_skip[currname] = TRUE
    new_processed_dataset = metabolomics_parsed_datasets[[currname]]
  }
  
  ##############################################################################
  # 10/15/2023 Update (Gayatri):
  # Adding some code to handle dataset with a single metabolite. This seems
  # to be causing issues while setting the rownames of the sample_data in 
  # metabolomics_processed_datasets. 
  
  if(nrow(metabolomics_parsed_datasets[[currname]]$sample_data)==1){
    print("--- Dataset has 1 metabolite; converting rownames")
    
    rownames(new_processed_dataset$sample_data)=new_processed_dataset$sample_data[,1]
    new_processed_dataset$sample_data=new_processed_dataset$sample_data[,-1]
    
    rownames(new_processed_dataset$control_data)=new_processed_dataset$control_data[,1]
    new_processed_dataset$control_data = new_processed_dataset$control_data[,-1]
  }
  ##############################################################################
  
  # Add BIDs from sample IDs
  new_processed_dataset$sample_meta$BID = substr(new_processed_dataset$sample_meta$sample_id,start = 1,stop = 5)
  
  # Check if the current vial labels are in the phenotypic data
  missing_samples = setdiff(colnames(new_processed_dataset$sample_data),
                            as.character(pheno_data$pheno_data$vialLabel))
  missing_samples_to_include = NULL
  
  # If ALL samples are missing, code will error and start next dataset
  if(length(missing_samples) == ncol(new_processed_dataset$sample_data)){
    print(paste("Skipping dataset",currname,"all vial labels missing from phenotypic data"))
    print("here is an example: ")
    print(colnames(new_processed_dataset$sample_data)[sample(which(missing_samples))[1]])
    next
  }
  
  # If there are sample labels that are not present in phenotype vial label data, 
  # samples are filtered out that have neither matching vial labels nor BIDs
  if(length(missing_samples) > 0 ){
    print(paste("-- In dataset", currname, length(missing_samples),"out of", length(colnames(new_processed_dataset$sample_data)), "samples are missing vial label IDs in the phenotype list"))
    missing_samples_bids = substr(missing_samples,start = 1,stop = 5)
    
    # Filter for missing samples to look at later (BID is present in phenotype data)
    missing_samples_to_include = missing_samples[missing_samples_bids %in% pheno_data$pheno_data$BID]
    if(length(missing_samples_to_include) > 0){
      print(paste("--- WARNING: Of missing samples,", length(missing_samples_to_include), "have BIDs in phenotype data"))  
    }
    
    # Filter for missing samples to remove (BID is not present in phenotype data)
    missing_samples_to_rem = missing_samples[!(missing_samples_bids %in% pheno_data$pheno_data$BID)]
    print(paste("--- Of missing samples,", length(missing_samples_to_rem), "do not have BIDs in phenotype data"))
    
    # Subset working data frame to only keep columns that are not in the list of samples to remove
    remaining_samples = setdiff(colnames(new_processed_dataset$sample_data),missing_samples_to_rem)
    new_processed_dataset$sample_data = new_processed_dataset$sample_data[,remaining_samples]
  }
  
  # Define tissue, platform (simplified), and site name (simplified) for storage in new
  # dataset name (newname). Variables also stored later in metabolomics_processed_datasets
  curr_tissue = strsplit(currname,split = ",")[[1]][1]
  curr_platform = strsplit(currname,split = ",")[[1]][2]
  curr_site = tolower(unique(new_processed_dataset$sample_meta$site))
  split_site = strsplit(curr_site, split = " ")[[1]]
  short_site_name = ifelse(length(split_site) > 0, ifelse(split_site[1] == "university", split_site[length(split_site)], split_site[1]), "")
  newname = paste(curr_tissue, curr_platform, short_site_name, sep = ",")
  
  # Define metadata (curr_meta) as the viallabel_data for which we have vial labels or BIDs
  # curr_meta is overwritten if there are missing samples to include which have BIDs that can be found in parent-daughter data
  if(newname %in% names(metabolomics_processed_datasets)){next}
  # curr_meta = pheno_data$viallabel_data[colnames(new_processed_dataset$sample_data),]
  curr_meta = pheno_data$pheno_data %>% filter(vialLabel %in% colnames(new_processed_dataset$sample_data))
  parent_daughter_biospec_meta = NULL
  
  # Add the sample order, type, raw peak intensity, and NA data from sample data and meta file to curr_meta file
  curr_meta$sample_order = new_processed_dataset$sample_meta[rownames(curr_meta),"sample_order"]
  curr_meta$sample_type = new_processed_dataset$sample_meta[rownames(curr_meta),"sample_type"]
  curr_meta$raw_intensity_pre = log2(apply(new_processed_dataset$sample_data[,rownames(curr_meta)],2,mean,na.rm=T)+1)
  
  # Assign object (curr_data) to new sample data which has been merged (named and unnamed) and 
  # is assigned phenotype data from vialdata and parent-daughter data
  curr_data = new_processed_dataset$sample_data
  
  # Convert values that are equal to or less than zero to NA
  if(any(curr_data == 0, na.rm = TRUE)){
    cat("\n\n---- THERE ARE 0 VALUES IN THIS DATA FRAME. Coverted to NA\n\n")
    curr_data[curr_data == 0] = NA
  }
  
  if(any(curr_data < 0, na.rm = TRUE)){
    cat("\n\n---- THERE ARE NEGATIVE VALUES IN THIS DATA FRAME. Coverted to NA\n\n")
    curr_data[curr_data < 0] = NA
  }
  
  # Sum the number of NAs per sample and add to meta data
  curr_meta$num_NAs = colSums(is.na(curr_data[,rownames(curr_meta)]))
  
  # Create the entry of the new dataset and add the dataset-level information
  metabolomics_processed_datasets[[newname]] = list()
  metabolomics_processed_datasets[[newname]]$tissue = curr_tissue
  metabolomics_processed_datasets[[newname]]$site = curr_site
  metabolomics_processed_datasets[[newname]]$platform = curr_platform
  metabolomics_processed_datasets[[newname]]$is_targeted = grepl("metab-t",curr_platform)
  
  # Add to parent_daughter_biospec_meta as new list item
  ## DELETE THIS:
  if(!is.null(parent_daughter_biospec_meta)) {
    metabolomics_processed_datasets[[newname]]$parent_daughter_biospec_meta = parent_daughter_biospec_meta
  }
  
  # Add the control level information directly from the original (merged) dataset
  metabolomics_processed_datasets[[newname]]$control_data = new_processed_dataset$control_data
  metabolomics_processed_datasets[[newname]]$control_sample_type = 
    new_processed_dataset$sample_meta[new_processed_dataset$sample_meta$sample_id %in% 
                                        colnames(new_processed_dataset$control_data),]
  
  # If only a single metabolite is present in dataset, a list of the different normalized 
  # versions of the dataset and the annotation data is created here.
  # This skips the downstream normalization step for datasets with multiple metabolites
  if(nrow(curr_data) == 1 || is.null(dim(curr_data))){
    print(paste0("-- In dataset ",currname, ", dataset has a single matabolite: ", rownames(curr_data)))
    message(paste0("-- No NA imputation for ",currname, " because dataset has 1 metabolite"))
    
    metabolomics_processed_datasets[[newname]]$normalized_data = list(
      "raw_imputed" = curr_data,
      "raw" = curr_data,
      "log2_imputed" = log2(1 + curr_data)
    )
    
    new_processed_dataset$row_annot$num_NAs = 
      rowSums(is.na(curr_data[rownames(new_processed_dataset$row_annot),]))
    
    metabolomics_processed_datasets[[newname]]$row_annot = new_processed_dataset$row_annot
    
    
    #### Update 10/17/2023: Gayatri Iyer
    #### The following 'next' statement throws the following error when it 
    #### encounters datasets with 1 metabolite:
    #### "Error: no loop for break/next, jumping to top level"
    #### Therefore, metabolomics_processed_datasets remains incomplete for
    #### this dataset and the sample_meta never gets allocated. Current work
    #### around is to remove the 'next' statement. This appears to fix the problem.
    #### May not be the best permanent fix though.
    
    # next
  }
  
  # Create logical vector indicating rows to remove
  # These rows have zero variance, no annotation, or rows with too many NAs (per max allowed NA rate)
  rows_to_remove = !(apply(curr_data, 1, sd, na.rm = T) > 0) | rownames(curr_data) == "" |
    rownames(curr_data) == "-" | rownames(curr_data) == "_" | 
    is.na(rownames(new_processed_dataset$sample_data))
  rows_to_remove[is.na(rows_to_remove)] = T
  row_percent_na = rowSums(is.na(curr_data)) / ncol(curr_data)
  rows_to_remove = rows_to_remove | row_percent_na > max_allowed_NA_rate
  
  # Update new sample data to assign new annotation data and control data with rows removed
  # Add num_NAs column to row_annot 
  curr_data = curr_data[!rows_to_remove,]
  curr_data_row_annot = new_processed_dataset$row_annot[!rows_to_remove,]
  curr_data_row_annot$num_NAs = rowSums(is.na(curr_data[rownames(curr_data_row_annot),]))
  
  metabolomics_processed_datasets[[newname]]$control_data = new_processed_dataset$control_data[!rows_to_remove,]
  
  # If NA values are present, impute these values using KNN
  #curr_data_imp = curr_data
  #na_data = list()
  if(any(is.na(curr_data))){
    print("-- RUNNING NA IMPUTATION")
    
    # Capture rows and columns that contain 1 NA
    if(sum(is.na(curr_data)) == 1) {
      na_per_metabolite = setNames(colnames(curr_data)[colSums(is.na(curr_data)) > 0],
                                   rownames(curr_data)[rowSums(is.na(curr_data)) > 0])
      na_per_sample = setNames(rownames(curr_data)[rowSums(is.na(curr_data)) > 0],
                               colnames(curr_data)[colSums(is.na(curr_data)) > 0])
      
      # Add NA identifying data to processed dataset
      metabolomics_processed_datasets[[newname]]$na_data = 
        list("by_metabolite" = na_per_metabolite, "by_sample" = na_per_sample)
    }
    
    # Capture rows and columns that contain > 1 NAs 
    if(sum(is.na(curr_data)) > 1) {
      na_data_temp = curr_data[!complete.cases(curr_data),colSums(is.na(curr_data)) > 0]
      
      # Capture the samples that are missing for each metabolite - CHECK
      na_per_metabolite <- vector(mode = "list", length = nrow(na_data_temp))
      names(na_per_metabolite) = rownames(na_data_temp)
      for(metabolite in names(na_per_metabolite)) {
        na_per_metabolite[[metabolite]] = colnames(na_data_temp[which(is.na(na_data_temp[metabolite,]))])
      }
      
      # Capture the metabolites that are missing for each sample - CHECK
      na_per_sample <- vector(mode = "list", length = ncol(na_data_temp))
      names(na_per_sample) = colnames(na_data_temp)
      for(labelid in names(na_per_sample)) {
        na_per_sample[[labelid]] = rownames(na_data_temp)[which(is.na(na_data_temp[,labelid]))]
      }
    }
    
    # Add NA identifying data to processed dataset
    metabolomics_processed_datasets[[newname]]$na_data = 
      list("by_metabolite" = na_per_metabolite, "by_sample" = na_per_sample)
    
    print(paste("--- Found", 
                sum(is.na(curr_data)), 
                "NAs in dataset out of", 
                sum(!is.na(curr_data)), 
                "in", 
                currname, 
                "-- running imputation"))
    
    curr_data_imp = min_val_knn_hybrid_imputation(curr_data, imputation_alpha, ml_method = "knn",
                                                  num_cores = min(15,nrow(curr_data)))
    
    if(max(abs(curr_data_imp-curr_data), na.rm = T) > 1e-05) {
      print(paste("ERROR: imputation changed non-NA numbers, check dataset:", currname))
      stop("Imputation changed non-NA numbers. Revise")
    }
    
  } else {
    print(paste("--- No NA imputation for",currname,"because data has no missing values"))
    curr_data_imp = curr_data
  }
  
  # Create log2 normalized version of the imputed data
  log2_imputed = log2(1 + curr_data_imp)
  
  # Re-calculated mean sample intensities, and add sample-level metadata to new entry
  curr_meta$raw_intensity_post = log2(apply(curr_data_imp[,rownames(curr_meta)],2,mean,na.rm=T)+1)
  metabolomics_processed_datasets[[newname]]$sample_meta = curr_meta
  
  # Apply median-MAD normalization to log2 normalized data for untargeted datasets only
  # This function is applied to each *sample* to account for sample-to-sample variation
  # Store the list of different normalized versions of the dataset to normalized_data object
  if(grepl("metab-u",currname)){
    median_mad_data = run_median_mad_norm(log2_imputed, zero_medians = F)
    metabolomics_processed_datasets[[newname]]$normalized_data = list(
      "raw_imputed" = curr_data_imp,
      "raw" = curr_data,
      "log2_imputed" = log2_imputed,
      "log2_median_mad" = median_mad_data
    )
    
    if(any(is.na(rownames(curr_data_imp)))){
      print(paste("NA row name in:",currname))
      break
    }
  } else {
    metabolomics_processed_datasets[[newname]]$normalized_data = list(
      "raw_imputed" = curr_data_imp,
      "raw" = curr_data,
      "log2_imputed" = log2_imputed
    )
  }
  
  # Enter the the annotation table of raw row names to final object
  metabolomics_processed_datasets[[newname]]$row_annot = curr_data_row_annot
  metabolomics_processed_datasets[[newname]]$rows_to_remove = rows_to_remove
  
  # Stored information about removed or imputed data
  percent_features_removed = rbind(percent_features_removed,
                                   c(newname,1-(nrow(curr_data)/length(row_percent_na))))
  percent_remaining_values_imputed = rbind(percent_remaining_values_imputed,
                                           c(newname,sum(is.na(curr_data))/(dim(curr_data)[1]*dim(curr_data)[2])))
  
  
  # Release some memory
  suppressWarnings({rm(curr_data); rm(curr_data_imp); rm(log2_imputed); rm(median_mad_data); gc()})
  print(paste0("++++++++++ FINISHED ",currname," ++++++++++++++++++++"))
  i = i + 1
}

# Final adjustments
percent_features_removed <- as.data.frame(percent_features_removed)
colnames(percent_features_removed) <- c("dataset", "percent_removed")
percent_remaining_values_imputed <- as.data.frame(percent_remaining_values_imputed)
colnames(percent_remaining_values_imputed) <- c("dataset", "percent_imputed")

# Ensure columns are numeric
percent_features_removed <- percent_features_removed %>%
  dplyr::mutate(percent_removed = as.numeric(percent_removed))

percent_remaining_values_imputed <- percent_remaining_values_imputed %>%
  dplyr::mutate(percent_imputed = as.numeric(percent_imputed))

cat("\n\nTOTAL NUMBER OF DATASETS PROCESSED: ", i)
cat("Total number of datasets with some features removed: ", sum(percent_features_removed[,2] > 0))
cat("Total number of datasets with some features imputed: ", sum(percent_remaining_values_imputed[,2] > 0))

```

### Plot percent of features removed and percent of remaining features imputed

These plots give an overview of data completeness and imputation levels across the different datasets. The first plot illustrates the percentage of features removed from each dataset due to excessive missingness. The second plot shows the percentage of remaining values that were imputed using the KNN method. This helps evaluate the degree of imputation performed, which is indicative of how much data completion was necessary after the initial removal.

```{r}

plot_data <- percent_features_removed %>%
  full_join(percent_remaining_values_imputed, by = "dataset") %>%
  rowwise() %>%
  mutate(platform = strsplit(dataset, "\\,")[[1]][2]) %>%
  mutate(tissue = strsplit(dataset, "\\,")[[1]][1]) 

# Plot missingness data
percent_plot <- ggplot(plot_data) +
  geom_col(aes(x = dataset, y = percent_removed*100, fill=platform)) +
  theme_bw() + theme(axis.text.x = element_text(size = 11)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) +
  xlab("") +
  ylab("%") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 100)) +  # Set y-axis limits
  ggtitle(paste0("Percent of features removed (>20% missing)"))
print(percent_plot)

percent_plot <- ggplot(plot_data) +
  geom_col(aes(x = dataset, y = percent_imputed*100, fill=platform)) +
  theme_bw() + theme(axis.text.x = element_text(size = 11)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) +
  xlab("") +
  ylab("%") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 100)) +  # Set y-axis limits
  ggtitle(paste0("Percent of remaining values KNN-imputed"))
print(percent_plot)

```

## Sanity checks

### Median sample intensity

This is a very basic test: we plot the data sample intensities (boxplots) next to the intensities of the reference and control samples of the same dataset. Only "flagged" datasets are plotted.

For PASS1B we did this:

* For all datasets we compare the median intensities of the groups (i.e., sex, week of training, controls are denoted as "0w") using the Kruskal-Wallis test, run for each sex separately
* Plot the boxplots for datasets with high inta-group (sex+week) heterogeneity, currently set to difference in log2 of the median intensity of at least $0.75$, for at least a single pair of samples from the same group
* Add boxplots of the reference/control samples

For Human Precovid we make these modifications:

*For all datasets we compare the median intensities of the groups (i.e., sex, timepoint, and intervention) using the Kruskal-Wallis test, run for each sex separately - moved to separate section with sample median decision making*

* Plot the boxplots for datasets with high intra-group (sex+week) heterogeneity, currently set to difference in log2 of the median intensity of at least $0.75$, for at least a single pair of samples from the same group
* Add boxplots of the reference/control samples
* For human precovid data, N is much larger, so though the median intensity cut off is still checked for the full tissue-method, the results are displayed per intervention-timepoint

```{r median sample intensity,out.height='50%',out.width='50%',message=FALSE,warning=FALSE,eval=T}

top2outliers = c()

for(currname in names(metabolomics_processed_datasets)){
  if(!grepl("metab-u",currname)){next}
  cat("------- START ",currname," -------\n")
  
  curr_data = metabolomics_processed_datasets[[currname]]
  
  ### GI update: 11/06/2023
  # Add 'vialLabel' as a column to sample_meta when missing #
  if (!("vialLabel" %in% colnames(curr_data$sample_meta))) {
    stop("Dataset ", currname, " does not have vialLabel column")
    cat(paste("Dataset ", currname, " does not have vialLabel column. 
              Converting rownames to vialLabel column \n"))
    curr_data$sample_meta <- 
      rownames_to_column(curr_data$sample_meta, 
                         var = "vialLabel")
  }
  ### End GI update ###
  
  curr_log2_data = curr_data$normalized_data$log2_imputed
  
  # Extract timepoints, randomization group, and sex to create an ordering variable and groups 
  curr_time = curr_data$sample_meta$Timepoint
  curr_rand = curr_data$sample_meta$randomGroupCode
  curr_sex = as.character(curr_data$sample_meta$Sex)
  
  ord = order(curr_sex, curr_time, curr_rand)
  
  curr_groups = paste(curr_time, curr_sex, curr_rand)
  names(curr_groups) = rownames(curr_data$sample_meta)
  
  cols = rainbow(length(unique(curr_groups)))
  names(cols) = unique(sort(curr_groups))
  
  # Calculate medians and IQRs for each sample 
  medians = apply(curr_log2_data, 2, median, na.rm = T)
  iqrs = colIQRs(as.matrix(curr_log2_data), na.rm = T)
  
  # Calculate the median difference between the max and min value within each 
  # Timepoint:sex:randomization group, and extract that maximum of the differences
  within_diffs = tapply(medians, curr_groups, function(x) max(x)-min(x))
  max_diff = max(within_diffs) 
  
  # Check for outliers by extracting residuals from linear model
  lm_model = lm(log10(medians) ~ curr_groups)
  lm_model_summary = summary(lm_model)
  lm_residuals = sort(abs(residuals(lm_model_summary)), decreasing = T)
  
  # Proceed to plotting figures if there are median differences within groups 
  # that are greater than or equal to 0.75
  if(max_diff < 0.75){next}
  
  cat("-- plotting figures if there are median differences within groups that are greater than or equal to 0.75")
  
  # Extract information on the top two 2 outliers in current data
  # Median values and IQRs are appended to top2outliers for each dataset
  currdf = data.frame(med = medians[ord], iqr = iqrs[ord],
                      group = curr_groups[ord], dataset = currname)
  top2outliers = rbind(top2outliers, currdf[names(lm_residuals)[1:2],])
  
  # For each Timepoint and each intervention group, make box plot
  for(tp in levels(curr_time)){
    for(interv in unique(curr_rand)){
      
      if(is.na(interv)) next
      # Filter for vial label names for the current Timepoint and intervention group
      active_samples = curr_data$sample_meta %>% 
        filter(Timepoint == tp, randomGroupCode == interv) %>% 
        pull(vialLabel) %>% as.character(.)
      
      if(length(active_samples) == 0){next}
      
      # Filter data for current Timepoint:intervention group samples only 
      curr_data_working = curr_log2_data[,active_samples] 
      curr_groups_working = curr_groups[active_samples]
      ord_working = order(intersect(ord[curr_data$sample_meta$Timepoint == tp], 
                                    ord[curr_data$sample_meta$randomGroupCode == interv]))
      
      # Set up graphical parameters and make a boxplot object bx to assign y-limits
      colnames(curr_data_working) = NULL
      par(cex.axis=0.8, mar=c(4, 3, 5, 2), xpd=TRUE)
      bx = boxplot(curr_data_working[,ord_working],plot=F)
      
      # Create boxplot with 
      boxplot(curr_data_working[,ord_working],
              main=paste0(currname,", ",tp," ",interv,"\nLog2 raw intensities (",nrow(curr_data_working)," metabolites)",
                          "\nMax within group log2FC: ",round(max_diff,digits = 2)),
              ylab = "log2 intensities",xaxt="n",col=cols[curr_groups_working[ord_working]],
              ylim = c(min(bx$stats[1,])-5, max(bx$stats[5,])+8),
              cex=0.4, cex.main=0.8)
      legend("bottom", legend = unique(names(cols[curr_groups_working[ord_working]])),
             fill = unique(cols[curr_groups_working[ord_working]]),
             ncol = 1,
             inset = c(0,-0.2))
    }}
  
  # Create log2 reference (control) data frame and annotation
  ref_data = log2(1 + curr_data$control_data)
  ref_data_annot = curr_data$control_sample_type$sample_type
  names(ref_data_annot) = curr_data$control_sample_type$sample_id
  ref_data = ref_data[,intersect(colnames(ref_data),names(ref_data_annot))]
  ref_data_annot = ref_data_annot[intersect(colnames(ref_data),names(ref_data_annot))]
  if(length(ref_data_annot) < 2){
    print("Not enough control samples to make boxplot")}
  
  # Order the data correctly
  ref_data = ref_data[,order(ref_data_annot)]
  ref_data_annot = ref_data_annot[order(ref_data_annot)]
  refcols = topo.colors(length(unique(ref_data_annot)))
  names(refcols) = unique(sort(ref_data_annot))
  
  # Plot control boxplot
  boxplot(ref_data,
          main=paste0(currname,"\nLog2 raw intensities of control samples"),
          ylab = "log2 intensities",xaxt="n",col=refcols[ref_data_annot],
          cex=0.4,cex.main=1)
  legend("top",legend = names(refcols), fill=refcols, ncol=2)
}

# Plot outliers as a table
top2outliers$vial = rownames(top2outliers)
kable(top2outliers,longtable=T,
      caption = "Top two outliers per dataset")  %>%
  kable_styling(latex_options = c("repeat_header"),font_size = 8)

```


### Kruskal-Wallis test report for sample-centering decision making

Here, we perform the following to determine whether or not the samples in a dataset should be sample-centered using median-MAD normalization:

1. Filter for untargeted datasets only
2. Compute the sample-wise median and quantile range
3. For each phase, compute the Kruskal-Wallis p-value between sexes and between exercise groups within each sex
4. Append p-values to the `kw_intensity_report`, determine the minimum p-value of these tests for each dataset, and assign sample-centering to any untargeted dataset that *does not* contain a comparison with a $p < 0.01$ 
5. Plot histograms for each comparison

```{r}

kw_intensity_report = c()

for(currname in names(metabolomics_processed_datasets)){
  
  if(!grepl("metab-u",currname)){next}
  
  # Parse imputed raw data and calculate the sample-wise median intensity and IQR
  curr_data = metabolomics_processed_datasets[[currname]]
  test_ss = data.frame(vialLabel = colnames(curr_data$normalized_data$raw_imputed),
                       samp.med = apply(curr_data$normalized_data$raw_imputed, 2, median, na.rm=T),
                       quant.75 = apply(curr_data$normalized_data$raw_imputed, 2, function(x) quantile(x, 0.75, na.rm=T)))
  
  # Combine median and IQR data with meta data for KW tests
  curr_int_data = metabolomics_processed_datasets[[currname]]$sample_meta %>%
    select(all_of(required_meta_vars)) %>%
    rownames_to_column("vialLabel") %>%
    right_join(., test_ss, by = "vialLabel")
  
  # Calculate the KW p-value between sexes for all exercise groups
  # Using if statement because ovaries and testes only have one sex
  if(length(unique(curr_int_data$Sex))>1){
    sex_pval_1 = kruskal.test(samp.med ~ Sex, data = curr_int_data)$p.value
    sex_pval_2 = kruskal.test(quant.75 ~ Sex, data = curr_int_data)$p.value
  } else {
    sex_pval_1 = NA_real_
    sex_pval_2 = NA_real_
  }
  
  # Save p-value data with other identifying information
  sex_intensity_report_temp = data.frame(dataset = currname,
                                         group = 'Sex',
                                         outcome = c('samp.med','quant.75'),
                                         p.value = c(sex_pval_1, sex_pval_2))
  
  # We now collect the KW p-values between exercise groups within each Sex
  group_intensity_report_temp = c()
  for(s in unique(curr_int_data$Sex)){
    curr_s_data = curr_int_data %>%
      filter(Sex == s)
    group_pval_1 = kruskal.test(samp.med ~ randomGroupCode, data = curr_s_data)$p.value
    group_pval_2 = kruskal.test(quant.75 ~ randomGroupCode, data = curr_s_data)$p.value
    
    # Save p-value data with other identifying information
    group_intensity_report_temp = rbind(group_intensity_report_temp, 
                                        data.frame(dataset = currname,
                                                   group = sprintf("randomGroupCode,%s",s),
                                                   outcome = c('samp.med','quant.75'),
                                                   p.value = c(group_pval_1, group_pval_2)))
  }
  
  # Append to report all information for each dataset
  kw_intensity_report = rbind(kw_intensity_report,
                              sex_intensity_report_temp,
                              group_intensity_report_temp)
  
}

# Make a dataframe that will delineate via "sample_ctr" whether or not that 
# dataset requires use of the sample-centered version
sample_centering_decision = kw_intensity_report %>%
  group_by(dataset) %>%
  slice_min(p.value) %>%
  mutate(sample_ctr = ifelse(p.value > 0.01, 1, 0))

# Plot a histogram of the Kruskal-Wallis p-values for each association
for(group_type in unique(kw_intensity_report$group)) {
  for(test_type in unique(kw_intensity_report$outcome)){
    
    hist_plot = kw_intensity_report %>% 
      na.omit(.) %>%
      filter(group == group_type, outcome == test_type)
    
    h <- ggplot(hist_plot, aes(x = p.value)) +
      geom_histogram(position = "identity") +
      xlab("P-values") +
      theme_bw() +
      ggtitle(paste0("Kruskal-Wallis test ", test_type," - ", group_type, " association"))
    print(h)
    
  }
}
```

#### Assign dataframe to use for each dataset

This section organizes and prepares metabolomics data for DEA by performing the following:

1. Read KW test results to decide whether untargeted data used should be sample-centered or not.
2. Select appropriate tables for DEA and assign to `data_use`:

* Metabolomics targeted (>12 features): KNN imputed, log2 transformed
* Metabolomics targeted (<=12 features): half-minimum imputation, log2 transformed
* Metabolomics untargeted sample-centered: KNN imputed, log2 transformed, median-MAD sample centered
* Metabolomics untargeted non-centered: KNN imputed, log2 transformed

```{r}

# Define the datasets that need to be sample-centered
# These datasets are untargeted and do not have Sex or exercise group associated
# with median intensity or median quartile
sample_center_datasets = sample_centering_decision %>%
  group_by(dataset) %>% 
  slice_min(sample_ctr) %>% 
  slice_head(n = 1) %>% 
  filter(sample_ctr == 1) %>%
  pull(dataset)

for (currname in names(metabolomics_processed_datasets)){
  
  curr_norm_data = metabolomics_processed_datasets[[currname]]$normalized_data
  
  # Targeted data
  if(grepl("metab-t",currname)){
    
    # If more than 12 metabolites, use log2 KNN imputed data
    if(nrow(curr_norm_data$raw) > 12){
      curr_data_use = curr_norm_data$log2_imputed
      print(paste0(currname, " is targeted, and has ", 
                   nrow(curr_norm_data$raw), 
                   " features, so we will use log2 imputed data"))
      
      # Otherwise perform half minimum imputation
    } else {
      curr_data_use <- curr_norm_data$raw
      for (i in 1:nrow(curr_data_use)) {
        if(sum(is.na(curr_data_use[i,]))!=0) {
          curr_data_use[is.na(curr_data_use)] <- 
            min(curr_data_use[i,],na.rm = T)/2
        }
      }
      curr_data_use = log2(1 + curr_data_use)
      print(paste0(currname, 
                   " is targeted, and has ", 
                   nrow(curr_norm_data$raw), 
                   " features, so we will perform imputation of the raw data using half-minimum imputation method and use log2 of the imputed data"))
    }
    
    # Untargeted data
  } else if (grepl("metab-u",currname)){
    
    # If dataset is annotated for sample-centering, use log2 imputed median-MAD
    if(currname %in% sample_center_datasets){
      curr_data_use = curr_norm_data$log2_median_mad
      print(paste0(currname, " is untargeted, and per KW tests, will be sample-centered"))
      
      # Otherwise use log2 imputed 
    } else {
      curr_data_use = curr_norm_data$log2_imputed
      print(paste0(currname, " is untargeted, and per KW tests, will NOT be sample-centered"))
    }
    
  } else {
    print("Error: Dataset not listed as targeted or untargeted")
    next
  }
  
  # Create the entry of the new dataset and add the data to use
  metabolomics_processed_datasets[[currname]]$data_use = curr_data_use
  
}
```

### Analysis of reference samples

We check the correlation between all reference and control samples by group. 
For example, if a dataset has both QC-Blank samples and QC-Reference samples, 
then we compute the overall correlation between each pair of *samples of the same group* 
and report the mean and min correlations. Samples and features with $>20\%$ of data missing are removed, 
and the missing samples that remain are imputed using KNN. 
The summary table shows the cases with a mean correlation of $<80\%$. 


```{r analyze qc samples, eval=FALSE, include=FALSE}

ref_corrs = list()
mean_intensity = list()
ref_sample_report = c()

# Clean control data and populate correlation data list (ref_corrs) and
# mean intensity data list (mean_intensity) by dataset and by QC sample type
for(currname in names(metabolomics_parsed_datasets)){
  
  mean_intensity[[currname]] = list()
  ref_corrs[[currname]] = list()
  
  # Parse data and recast as numeric
  curr_data = metabolomics_parsed_datasets[[currname]]
  
  if (length(curr_data)<=3) {next}
  
  curr_ctrl_data = curr_data$control_data
  curr_ctrl_data = recast_numeric_data_frame(curr_ctrl_data)
  if(is.null(dim(curr_ctrl_data)) || length(curr_ctrl_data) < 2){next}
  
  # Remove samples and features with >20% missing data
  curr_ctrl_data = curr_ctrl_data[,colSums(is.na(curr_ctrl_data))/nrow(curr_ctrl_data) < 0.2]
  curr_ctrl_data = curr_ctrl_data[rowSums(is.na(curr_ctrl_data))/ncol(curr_ctrl_data) < 0.2,]
  if(is.null(dim(curr_ctrl_data)) || length(curr_ctrl_data) < 2){next}
  
  # Remaining missing samples are imputed using KNN with capture.output to avoid printing 
  capture.output(
    {curr_ctrl_data = impute::impute.knn(as.matrix(curr_ctrl_data))$data}, file = NULL)
  
  # Remove samples with zero variance
  curr_ctrl_data = curr_ctrl_data[, apply(curr_ctrl_data, 2, sd) > 0]
  if(is.null(dim(curr_ctrl_data)) || length(curr_ctrl_data) < 2){next}
  
  # Isolate control sample IDs and the type of control data 
  qc_types = curr_data$sample_meta %>%
    filter(sample_id %in% colnames(curr_ctrl_data)) %>% 
    pull(sample_type,sample_id)
  
  # Perform Spearman correlation on control data
  ctrl_corr_data = cor(curr_ctrl_data, method = "spearman")
  
  # Filter for samples of specific QC-type and store correlation and mean intensity data
  for(qc_type in unique(qc_types)){
    
    # Parse correlation data for specific dataset and QC type
    inds = names(qc_types)[qc_types == qc_type]
    inds_corr_data = ctrl_corr_data[inds,inds]
    ref_corrs[[currname]][[qc_type]] = inds_corr_data
    if(is.null(inds_corr_data) || length(inds_corr_data) == 1){next}
    
    # Parse mean intensity data for specific dataset and QC type
    curr_qc_data = as.matrix(curr_ctrl_data[,inds])
    curr_qc_data = apply(log2(1 + curr_qc_data), 2, mean, na.rm = T) 
    mean_intensity[[currname]][[qc_type]] = curr_qc_data
    
    # Filter for half of correlation triangle to avoid redundancy 
    inds_corr_data_tri = inds_corr_data[lower.tri(inds_corr_data)]
    
    # Calculate mean and minimum correlation coefficients, and mean-mean intensity
    # Append to the sample report
    v = c()
    v["mean_rho"] = format(mean(inds_corr_data_tri, na.rm = T), digits = 3)
    v["min_rho"] = format(min(inds_corr_data_tri, na.rm = T), digits = 3)
    v["mean_intensity"] = format(mean(curr_qc_data, na.rm = T), digits = 3)
    ref_sample_report = rbind(ref_sample_report, c(qc_type, v))
    rownames(ref_sample_report)[nrow(ref_sample_report)] = currname
  }
}

# Filter report for QC types with correlations less than 80%
kable(ref_sample_report[as.numeric(ref_sample_report[,"mean_rho"]) < 0.8,], longtable=T,
      caption = "Reference samples: Mean correlation < 0.8") %>%
  kable_styling(font_size = 8,latex_options = c("hold_position", "repeat_header"))

```

## PCA Analysis

### Visualization

PCA analyses are performed on the log2 imputed data. 
We plot PC1 and PC2 for each dataset, faceting by Sex and randomization group.  

```{r pca visualization,out.height='50%',out.width='50%',message=FALSE,warning=FALSE,eval=T}
for(currname in names(metabolomics_processed_datasets)){
  
  d = metabolomics_processed_datasets[[currname]]
  if(nrow(d$normalized_data[[1]]) < 5){next}
  
  # Isolate log2 imputed data
  curr_data = d$normalized_data[["log2_imputed"]]
  #curr_data =log2(1 + abs(min(d$normalized_data[["unnorm"]])) + d$normalized_data[["unnorm"]] ) #old
  # Select metabolites with SD greater than 0
  curr_data = curr_data[apply(curr_data, 1, sd) > 0,]
  
  # Run PCA and subset either the number of PCs specified in setup block,
  # or select the number of rows (features) in dataset if num_pcs
  curr_pca = prcomp(t(curr_data),scale. = T,center = T)
  curr_pcax = curr_pca$x[,1:min(num_pcs,nrow(curr_data))]
  
  # Parse the explained variance for each PC
  explained_var = summary(curr_pca)[["importance"]][2,1:ncol(curr_pcax)]
  
  # Append PCA data and variance to metabolomics_processed_datasets
  metabolomics_processed_datasets[[currname]][["pca"]] = list(
    pcax = curr_pcax, explained_var = explained_var
  )
  
  # Making dataframe for plotting with PCs and meta data of interest
  plot_df = data.frame(
    PC1 = curr_pcax[,1],PC2 = curr_pcax[,2],PC3 = curr_pcax[,3],PC4 = curr_pcax[,4],
    Timepoint = d[["sample_meta"]][,"Timepoint"],
    Sex = as.character(d[["sample_meta"]][,"Sex"]),
    stringsAsFactors = F,
    randomGroupCode = as.character(d[["sample_meta"]][,"randomGroupCode"])
  )
  
  # Identify Sex variables 
  plot_df$Sex[plot_df$Sex=="1"] = "female"
  plot_df$Sex[plot_df$Sex=="2"] = "male"
  
  # Plot faceted PCA and color by Timepoint
  p = ggplot(plot_df) + 
    geom_point(aes(x = PC1, y = PC2, color=Timepoint)) +
    xlab(paste0("PC1 (",round(explained_var["PC1"]*100, 1), " %)")) +
    ylab(paste0("PC2 (",round(explained_var["PC2"]*100, 1), " %)")) +
    facet_wrap(~ Sex + randomGroupCode) +
    scale_color_manual(values = acute_timepoint_cols) +
    ggtitle(currname)
  plot(p)
  
  # Plot all in PCA, colored by Sex
  q = ggplot(plot_df) + 
    geom_point(aes(x = PC1, y = PC2, color=Sex)) +
    xlab(paste0("PC1 (",round(explained_var["PC1"]*100, 1), " %)")) +
    ylab(paste0("PC2 (",round(explained_var["PC2"]*100, 1), " %)")) +
    scale_color_manual(values = sex_cols) +
    ggtitle(currname)
  plot(q)
  
  # Plot all in PCA, colored by randomization group
  r = ggplot(plot_df) + 
    geom_point(aes(x = PC1, y = PC2, color=randomGroupCode)) +
    xlab(paste0("PC1 (",round(explained_var["PC1"]*100, 1), " %)")) +
    ylab(paste0("PC2 (",round(explained_var["PC2"]*100, 1), " %)")) +
    scale_color_manual(values = exercise_type_cols) +
    ggtitle(currname)
  plot(r)
}

```

### Association analysis

Here we analyze the top principal components in each dataset and compute 
their association with the selected variables above. 
We use Spearman correlation and a linear test for significance. 

For each dataset we also add the correlations among the metadata variables. 
These two analyses should be interpreted together, as the analyzed variables are not independent.


```{r pca association analysis,eval=T,message=FALSE,warning=FALSE,results="hide"}
# Define technical and biological variables to be considered in the QC
pipeline_qc_cols = c("sample_order",
                     "raw_intensity_post",
                     "num_NAs")
biospec_cols = c("Sex",
                 "randomGroupCode",
                 "calculatedAge",
                 "race",
                 "Latinx",
                 "wtkg_pcaa",
                 "coll_staffid",
                 "staffID",
                 "codedsiteid",
                 "Timepoint")
```

```{r pca association analysis2,out.height='50%',out.width='50%',message=FALSE,warning=FALSE}
pcs_vs_qc_var_report = c()
meta_var_assoc_report = c()

for(currname in names(metabolomics_processed_datasets)) {
  
  d = metabolomics_processed_datasets[[currname]]
  if(nrow(d$normalized_data[[1]]) < 5){next}
  
  # Combine all unique qc and biological variables into one vector
  # that are also present in the sample_meta dataframe, remove text groups
  
  # TO DO: is there some way to make Timepoint categorical and compare multi-level categorical variable
  curr_meta_cols = union(biospec_cols, pipeline_qc_cols)
  curr_meta_cols = intersect(curr_meta_cols, names(d$sample_meta))
  curr_meta = d$sample_meta[,curr_meta_cols] %>% 
    mutate(randomgroupcodenum = as.numeric(factor(randomGroupCode)),
           timepointnum = as.numeric(factor(Timepoint))) %>%
    select_if(is.numeric) 
  
  # Remove metadata variables with NAs or zero variance
  curr_meta = curr_meta[,!apply(is.na(curr_meta), 2, any)]
  curr_meta = curr_meta[,apply(curr_meta, 2, sd) > 0]
  
  # Get the PCA data and variances
  curr_pcax = d$pca$pcax
  curr_variances = d$pca$explained_var
  
  # First, compute correlations between metadata and PCA data
  corrs = cor(curr_pcax, curr_meta, method="spearman")
  corrsp = pairwise_eval(curr_pcax, curr_meta, func = pairwise_association_wrapper, f = 1)
  
  # Make correlation plot between metadata and PCA data
  corr_plot = 
    ggcorrplot((corrs), lab=T, lab_size=2.5, hc.order = F) + 
    ggtitle(currname) +
    scale_x_discrete(breaks = rownames(corrs), 
                     labels = paste0(rownames(corrs), " (",round(curr_variances[rownames(corrs)]*100, 1), " %)")) + 
    theme(plot.title = element_text(hjust = 0.5, size=20))
  print(corr_plot)
  
  # Add data to the report
  # Only saves correlations with p-values below designated threshold p_thr
  for(pc in 1:nrow(corrsp)) {
    for(meta_var in 1:ncol(corrsp)) {
      if(corrsp[pc,meta_var] > p_thr) {next}
      pcs_vs_qc_var_report = rbind(pcs_vs_qc_var_report,
                                   c(currname,
                                     rownames(corrsp)[pc],
                                     colnames(corrsp)[meta_var],
                                     corrs[pc,meta_var],
                                     corrsp[pc,meta_var]))
    }
  }
  
  # Second, compute correlations between all metadata
  curr_meta = curr_meta[!apply(is.na(curr_meta), 1, any),]
  corrs = cor(curr_meta, method = "spearman")
  corrsp = pairwise_eval(curr_meta, func = pairwise_association_wrapper, f = 1)
  
  # Make correlation plot between metadata and PCA data
  print(ggcorrplot(corrs, lab=T, lab_size=2.5, hc.order = F) + ggtitle(currname))
  
  # Add data to the report. This loop only saves p-values that are below the
  # designated threshold p_thr and that are computed between bio specimen metadata
  # and pipeline QC metrics
  for(meta_var1 in rownames(corrsp)){
    for(meta_var2 in rownames(corrsp)){
      if(meta_var1 == meta_var2){break}
      if(meta_var1 %in% biospec_cols &&
         meta_var2 %in% biospec_cols) {next}
      if(corrsp[meta_var1,meta_var2] > p_thr) {next}
      meta_var_assoc_report = rbind(meta_var_assoc_report,
                                    c(currname,
                                      meta_var1,
                                      meta_var2,
                                      corrs[meta_var1,meta_var2],
                                      corrsp[meta_var1,meta_var2]))
    }
  }
  
}

# Format the reports for a nicer presentation in a table
pcs_vs_qc_var_report = data.frame(pcs_vs_qc_var_report) %>%
  mutate_at(c(4,5), .funs = ~format(as.numeric(.), digits = 3))
meta_var_assoc_report = data.frame(meta_var_assoc_report) %>%
  mutate_at(c(4,5), .funs = ~format(as.numeric(.), digits = 3))

# Set rownames for table output
if(!is.null(pcs_vs_qc_var_report)){
  colnames(pcs_vs_qc_var_report) = c("Dataset(tissue,site)","PC",
                                     "qc_metric","rho(spearman)","p-value")
}
if(!is.null(dim(meta_var_assoc_report))){
  colnames(meta_var_assoc_report) = c(
    "Dataset(tissue,site)","Var1","Var2","rho(spearman)","p-value")
}
```

Print the significant results in tables

```{r pca association analysis3,eval=T}
kable(pcs_vs_qc_var_report, longtable=T,
      caption = "Correlations between PCs and other variables")  %>%
  kable_styling(latex_options = c("repeat_header"), font_size = 8)

kable(meta_var_assoc_report, longtable=T,
      caption = "Correlations between metadata variables")%>%
  kable_styling(latex_options = c("repeat_header"), font_size = 8)
```

### Outliers

Here, PC-based outliers are flagged by examining the boxplot of each PC and extending its whiskers to the predefined multiplier above and below the interquartile range (*3x the IQR as pre-defined in this script and in the PASS1B landscape paper*). Samples outside this range are flagged. Note that a sample may be flagged in PC3, but here we only visualize in PCs 1 and 2. 

**Samples are flagged using an automatic analysis of the principal components. Such analyses may flag samples because of true biological effects and thus further examination is required before determining if flagged samples represent low quality samples. Note that we plot outliers for the log2 imputed data, and outliers may be more likely to appear in the raw data.**

```{r pca outliers,out.height='50%',out.width='50%'}
pca_outliers_report = c()

for(currname in names(metabolomics_processed_datasets)){
  d = metabolomics_processed_datasets[[currname]]
  if(nrow(d$normalized_data[[1]]) < 5) {next}
  
  # Get PCA values and variances for the pre-defined number of PCs 
  # delineated for outlier analysis
  curr_pcax = d$pca$pcax[,1:num_pcs_for_outlier_analysis]
  curr_pcavar = d$pca$explained_var[1:num_pcs_for_outlier_analysis]
  
  pca_outliers = c()
  
  # Get the outlier values using the IQR threshold delineated for each PC
  for(pc in colnames(curr_pcax)){
    outlier_values <- boxplot.stats(curr_pcax[,pc], coef = OUTLIER_IQR_THR)$out
    
    # Save individual outliers with their information in the report
    for(outlier in names(outlier_values)){
      pca_outliers_report = rbind(pca_outliers_report,
                                  c(currname, pc, curr_pcavar[pc], outlier,
                                    format(outlier_values[outlier], digits=5)))
      
      # Store outlier from individual PC in combined vector from all PCs
      # for outlier visualization
      if(!is.element(outlier, names(pca_outliers))){
        pca_outliers[outlier] = outlier_values[outlier]
      }
    }
  }
  
  # Plot the outliers
  if(length(pca_outliers) > 0) {
    
    # Add a column delineating which samples are outliers 
    plot_df = data.frame(curr_pcax) %>%
      mutate(outliers = rownames(curr_pcax) %in% names(pca_outliers)) %>%
      mutate(col = if_else(outliers == T, "green","black"))
    
    plot(plot_df$PC1, plot_df$PC2, pch = as.numeric(plot_df$outliers),
         col = plot_df$col, lwd = 2, cex = 1,
         main = paste(currname,"flagged outliers"),
         xlab = paste0("PC1 (",round(curr_pcavar["PC1"]*100, 1), " %)"), 
         ylab = paste0("PC2 (",round(curr_pcavar["PC2"]*100, 1), " %)"))
    
    pairs(plot_df[, c("PC1", "PC2", "PC3")], 
      pch = as.numeric(plot_df$outliers), 
      col = plot_df$col, 
      lwd = 2, cex = 1.2,
      main = paste(currname, "Flagged Outliers"))
    
  }
}

# Re-organize outliers report for sample report printing
colnames(pca_outliers_report) =  c("Dataset","PC","Variance","Sample","Score")

outlier_report = data.frame(pca_outliers_report) %>%
  select(Dataset, PC, Sample) %>%
  group_by(Dataset, Sample) %>%
  reframe(PC = paste(PC, collapse = ", "))

outlier_report <- outlier_report %>% filter(!(Dataset=="t02-plasma,metab-t-oxylipneg,emory" & Sample!="11386040214"))

split_names = stringr::str_split(outlier_report$Dataset, ",", simplify = TRUE)
split_ome_tissue = OME_TISSUE_CODE %>% select(tissue, tissue_code) %>% distinct()
writeout_outliers = data.frame(
  tissue_code = split_names[,1],
  ome = split_names[,2],
  source = split_names[,3],
  Dataset = outlier_report$Dataset,
  stringsAsFactors = FALSE
) %>%
  left_join(., split_ome_tissue, by = "tissue_code") %>%
  merge(., outlier_report, by = "Dataset") %>%
  distinct(Sample, Dataset, PC, .keep_all = TRUE) %>%
  select(Dataset,tissue, ome, Sample, PC)

# write.csv(writeout_outliers, file = "~/Downloads/outlier_report.csv", row.names = F)
distinct_combos = writeout_outliers %>% distinct(tissue, ome)
for(row in seq(nrow(distinct_combos))){
  single_combo = distinct_combos[row,]
  if(grepl("metab-t", single_combo$ome)) local_path = file.path(repo_local_dir, "data/tmp/metab_outliers/metabolomics-targeted/metadata/")
  if(grepl("metab-u", single_combo$ome)) local_path = file.path(repo_local_dir, "data/tmp/metab_outliers/metabolomics-untargeted/metadata/")
  outliers_for_combo = writeout_outliers %>% 
    filter(tissue == single_combo$tissue,
           ome == single_combo$ome)
  write_with_path_name(outliers_for_combo, 
                     local_path = local_path,
                     ome = single_combo$ome,
                     tissue = single_combo$tissue,
                     data_category = "metadata",
                     data_details = "removed-samples",
                     version = "1.3", 
                     return_name_only = FALSE)
}


kable(outlier_report, longtable = T,
      caption = "Outliers detected by tissue PCA data") %>%
  kable_styling(font_size = 8, latex_options = c("hold_position", "repeat_header"))

```



## CV calculations

In metabolomics, the coefficient of variation (CV) measures the relative variability of metabolite concentrations, calculated as the ratio of the standard deviation to the mean (expressed as a percentage). It is a key metric for assessing the reproducibility and reliability of measurements, especially in distinguishing technical variation from true biological differences. Low CVs indicate consistent and precise data, which is critical for both targeted and untargeted metabolomics workflows. For targeted analyses, CVs under 10-15% are considered excellent, while in untargeted studies, CVs below 20-30% are generally acceptable, reflecting the higher complexity and variability of untargeted methods.

These calculations are based on named metabolites only and using QC-Pooled samples for all datasets except for Duke, which uses QC-ExternalStandard samples.

```{r load data}
control_samples <- c()
for (i in names(metabolomics_processed_datasets)) {
  if (!(is.null(metabolomics_processed_datasets[[i]]$control_data))) {
    if ("is_named" %in% colnames(metabolomics_processed_datasets[[i]]$row_annot)) {
      message(i, " control data, is named")
      named_features <- metabolomics_processed_datasets[[i]]$row_annot %>%
        filter(is_named==TRUE) %>%
        select(metabolite_name) %>%
        pull()
      curr_dat <- t(metabolomics_processed_datasets[[i]]$control_data[named_features,]) %>%
        data.frame(check.names = F) %>%
        rownames_to_column(var="sample_id")
    } else {
      curr_dat <- t(metabolomics_processed_datasets[[i]]$control_data) %>%
        data.frame(check.names = F) %>%
        rownames_to_column(var="sample_id")
      message("ATTENTION! 'is_named' annotation MISSED in ", i)
    }
    curr_data_type <- metabolomics_processed_datasets[[i]]$control_sample_type
    curr_dat2 <- inner_join(curr_data_type, curr_dat, by="sample_id")
    control_samples[[i]] <- curr_dat2
  }else{
    message("WARNING: no data available for ", i)
  }
}
```


### QC sample types count

Count all the number of samples for each qc type

```{r qc sample count}
control_sample_types <- c()
for (i in names(control_samples)) {
  curr_dat <- unique(control_samples[[i]]$sample_type)
  control_sample_types[[i]] <- curr_dat
}

control_sample_type_count <- c()
for (i in names(control_samples)) {
  control_sample_type_count[[i]] <- table(control_samples[[i]]$sample_type)
}
control_sample_type_count2 <- bind_rows(control_sample_type_count, .id="dataset") %>%
  data.frame(check.names = F) %>%
  replace(is.na(.),0)

# Extract metadata from "dataset"
control_sample_type_count2 <- control_sample_type_count2 %>%
  mutate(
    tissue = sub(",.*", "", dataset), 
    assay = sub("^[^,]+,([^,]+),.*$", "\\1", dataset),
    site = ifelse(grepl("^[^,]+,[^,]+,$", dataset), NA, sub(".*,(.*)$", "\\1", dataset))
  )

# Add the Duke label
# control_sample_type_count2$site[is.na(control_sample_type_count2$site)] <- "duke"

control_sample_type_count2_long <- control_sample_type_count2 %>%
  pivot_longer(
    cols = starts_with("QC"),  
    names_to = "qc_type",   
    values_to = "qc_counts"
  )

# fix variable types
control_sample_type_count2_long <- control_sample_type_count2_long %>%
  dplyr::mutate(
    qc_type = as.character(qc_type),
    assay = as.character(assay),
    qc_counts = as.numeric(qc_counts),
    tissue_name = sub("^t\\d+-", "", tissue)
  )

```


### Plot Counts

```{r plot counts, fig.height=8, fig.width=8}

p1 <- ggplot(control_sample_type_count2_long, aes(x = qc_type, y = qc_counts, fill = assay)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ tissue, scales = "free_y") +
  labs(
    title = "Sample QC Counts by Assay and Tissue",
    x = "QC Type",
    y = "Counts",
    fill = "Assay"
  ) +
  theme_linedraw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

print(p1)


p2 <- control_sample_type_count2_long %>%
  filter(qc_type == "QC-Pooled", site != "duke") %>% 
  ggplot(aes(x = assay, y = qc_counts, fill = site)) +
  facet_wrap(~ tissue_name) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Sample count for QC-Pooled",
    caption = "QC-Pooled not available for Duke",
    x = "Assay",
    y = "QC Counts",
    fill = "Tissue"
  ) +
  theme_linedraw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) + coord_flip()

print(p2)

p3 <- control_sample_type_count2_long %>%
  filter(qc_type == "QC-ExternalStandard", site == "duke") %>% 
  ggplot(aes(x = assay, y = qc_counts, fill = assay)) +
  facet_wrap(~ tissue) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Sample count for QC-ExternalStandard",
    subtitle = "Only Duke",
    x = "Assay",
    y = "QC Counts",
    fill = "Tissue"
  ) +
  theme_linedraw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) + coord_flip()

print(p3)

```


### CV Calculations

Select QC-samples data used for calculating the CVs

```{r cv pre calculations}

qc_pooled_samples <- c()
for (a in names(control_samples)) {
  curr_qc_pooled <- control_samples[[a]] %>%
    filter(sample_type=="QC-Pooled") %>%
    select(-c(sample_type:BID))
  qc_pooled_samples[[a]] <- curr_qc_pooled
}

qc_exStd_samples <- c()
for (a in names(control_samples)) {
  curr_qc_ref <- control_samples[[a]] %>%
    filter(sample_type=="QC-ExternalStandard") %>%
    select(-c(sample_type:BID))
  qc_exStd_samples[[a]] <- curr_qc_ref
}

```

Calculate the CVs using a function

```{r cv function and calculation}
# Function to calculate the CVs
calculate_cv <- function(qc_samples, exclude_columns = 1) {
  
  # Ensure that qc_samples is a list
  if (!is.list(qc_samples)) {
    stop("Input qc_samples must be a list.")
  }
  
  # Use purrr to map over the list and bind rows into a data frame
  qc_cv_df <- purrr::imap_dfr(qc_samples, ~ {
    batch_name <- .y
    curr_dat <- .x
    
    # Check if .x is a data frame or matrix
    if (!is.data.frame(curr_dat) && !is.matrix(curr_dat)) {
      warning(paste("Element", batch_name, "is not a data frame or matrix. Skipping."))
      return(NULL)
    }
    
    # Ensure there are enough columns to process
    if (ncol(curr_dat) <= exclude_columns) {
      warning(paste("Element", batch_name, "has insufficient columns after excluding specified columns. Skipping."))
      return(NULL)
    }
    
    # Extract metabolite data
    metabolite_data <- curr_dat[, -exclude_columns, drop = FALSE]
    metabolite_names <- colnames(metabolite_data)
    
    # Calculate CVs
    cv_values <- vapply(seq_along(metabolite_data), function(i) {
      values <- metabolite_data[[i]]
      non_na_values <- values[!is.na(values)]
      
      # Check for sufficient data points
      if (length(non_na_values) < 2) {
        return(NA_real_)  # Not enough data to compute SD
      }
      
      mean_value <- mean(non_na_values)
      sd_value <- sd(non_na_values)
      
      # Check for zero or NA mean
      if (is.na(mean_value) || mean_value == 0) {
        return(NA_real_)
      } else {
        return((sd_value / mean_value) * 100)
      }
    }, FUN.VALUE = numeric(1))
    
    # Parse batch_name to extract batch information
    batch_info <- strsplit(batch_name, ",")[[1]]
    # Ensure batch_info has at least three elements
    batch_info <- c(batch_info, rep(NA_character_, 3 - length(batch_info)))
    tissue <- batch_info[1]
    assay <- batch_info[2]
    site <- batch_info[3]
    if (is.na(site) || site == "") {
      site <- "duke"
    }
    
    # Create a data frame for this batch with additional columns
    tibble(
      tissue_assay_sites = batch_name,
      tissue = tissue,
      assay = assay,
      site = site,
      feature_id = metabolite_names,
      feature_cv = cv_values
    )
  })
  
  return(qc_cv_df)
}

duke_assays <- c("t02-plasma,metab-t-conv,duke", 
                 "t02-plasma,metab-t-imm-crt,duke", 
                 "t10-muscle,metab-t-nuc,duke", 
                 "t11-adipose,metab-t-acoa,duke", 
                 "t11-adipose,metab-t-ka,duke", 
                 "t11-adipose,metab-t-nuc,duke")

# QC Pool samples based CV, excluding Duke
qc_pooled_samples_df <- calculate_cv(qc_samples = qc_pooled_samples[setdiff(names(qc_pooled_samples), duke_assays)])
# QC external Standards based CV, only for Duke
qc_exStd_cv_df <- calculate_cv(qc_samples = qc_exStd_samples[names(qc_pooled_samples) %in% duke_assays])

metab_cvs <- rbind(qc_pooled_samples_df, qc_exStd_cv_df)
```


### Plot CVs

For each site, plot the distribution of CVs (boxplot) for every assay and tissue


```{r}
cv_cutoff <- 30

for(sit in unique(metab_cvs$site)){
  fdf <- metab_cvs %>% filter(site==sit)
  p <- ggplot(fdf, aes(x = assay, y = feature_cv, fill = assay)) +
    geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.7, na.rm = TRUE) +
    geom_jitter(width = 0.1, size = 1, alpha = 0.6, aes(color = assay), na.rm = TRUE) +
    geom_hline(yintercept = cv_cutoff, linetype = "dashed", color = "red") +
    facet_wrap(~ tissue, scales = "free") +
    labs(
      title = paste0(toupper(sit), ": CV Distributions"),
      x = "Assay",
      y = "Feature CV (%)",
      fill = "Assay",
      color = "Assay"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(size = 12)
    ) +
    scale_y_continuous(limits = c(0, 100)) 
  
  print(p)
}

```

Plot the percentage of features over a certain CV threshold (30%) for each assay and tissue

```{r}

# Create a new variable indicating whether each feature_cv is above or below 30%
cv_summary <- metab_cvs %>%
  mutate(
    cv_category = ifelse(feature_cv > cv_cutoff, paste0("CV > ",cv_cutoff, "%"), paste0("CV <= ", cv_cutoff, "%"))
  ) %>%
  group_by(site, assay, tissue, cv_category) %>%
  summarise(
    count = n(),
    .groups = 'drop'
  ) %>%
  group_by(site, assay, tissue) %>%
  mutate(
    total = sum(count),
    percentage = (count / total) * 100
  )

# Create the stacked bar plot
for(sit in unique(cv_summary$site)){
  fdf <- cv_summary %>% filter(site==sit)
  p2 <- ggplot(fdf, aes(x = assay, y = percentage, fill = cv_category)) +
    geom_bar(stat = 'identity') +
    facet_grid(site ~ tissue, scales = "free_x") +
    labs(
      title = paste(toupper(sit)),
      x = "Assay",
      y = "Percentage of Features",
      fill = "CV Category"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  print(p2)
}

```

### Details: CVs for every metabolite in every assay

Includes the median CV for each assay and tissue, and truncates the bars at 100% CV (for better visualization)

```{r}

# Function to plot non-NA values for a specific assay with truncated bars
plot_assay_values <- function(ds, assay_name, max_limit = 100) {
  
  # Check if the assay exists in the list
  if (!assay_name %in% unique(ds$tissue_assay_sites)) {
    stop("Assay not found in the list.")
  }
  
  # Extract the data for the specific assay
  assay_data <- ds %>% filter(tissue_assay_sites == assay_name)
  
  # Remove NA values
  assay_data_nna <- na.omit(assay_data)
  
  # Calculate the median and add it to the data frame
  assay_data_nna_with_median <- assay_data_nna %>%
    group_by(tissue_assay_sites, tissue, assay, site) %>%
    summarise(
      feature_id = "median",
      feature_cv = median(feature_cv, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    bind_rows(assay_data_nna) %>% 
    arrange(tissue_assay_sites, tissue, assay, site)
  
  # Sort the data frame in descending order of CV
  assay_data_nna_with_median <- assay_data_nna_with_median[order(-assay_data_nna_with_median$feature_cv), ]
  
  # Add columns for truncation logic
  plot_data <- assay_data_nna_with_median %>%
    mutate(
      exceeded = feature_cv > max_limit,
      truncated_value = pmin(feature_cv, max_limit)
    )
  
  # Create a bar plot with truncated bars
  p <- ggplot(plot_data, 
              aes(x = reorder(feature_id, -feature_cv), 
                  y = truncated_value,
                  fill = feature_id == "median")) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("steelblue", "orange"), guide = FALSE) +
    # Add arrow heads for truncated bars
    geom_segment(
      data = filter(plot_data, exceeded),
      aes(x = reorder(feature_id, -feature_cv),
          xend = reorder(feature_id, -feature_cv),
          y = max_limit - 2,
          yend = max_limit),
      arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
      color = "black"
    ) +
    # Add original values as text for truncated bars
    geom_text(
      data = filter(plot_data, exceeded),
      aes(label = sprintf("%.1f%%", feature_cv),
          y = max_limit - 5),
      size = 2.5,
      hjust = 1
    ) +
    geom_hline(yintercept = 15, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 30, linetype = "dashed", color = "blue") +
    scale_y_continuous(
      limits = c(0, max_limit),
      expand = expansion(mult = c(0, 0.05))
    ) +
    labs(
      title = paste("Dataset:", assay_name),
      x = "Metabolite",
      y = "CV (%)"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 0, hjust = 1, vjust = 1, size = 8),
      axis.text.y = element_text(size = 6)
    ) + 
    coord_flip()
  
  return(p)
}

for(dataset in unique(metab_cvs$tissue_assay_sites)){
  p <- plot_assay_values(ds = metab_cvs, assay_name = dataset)
  print(p)
}

```


# QC Report Summary

The QC report illustrates the relative consistency of samples and assay controls,
calculates CVs across platforms to allow for analysts and others to visualize
the dataset, identify which samples should be included vs excluded, decide
which datasets should be sample centered vs platform centered, etc. 

Based on the decisions from this report, new datasets are generated in 
`Metab_outlier_report`. 



