---
title: "QC Normalization: Generation and Visualization"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(edgeR)
library(dplyr)
library(here)
library(MotrpacHumanPreSuspensionData)

devtools::load_all()
```

## Overview

This document walks through the QC-normalization generation process across all supported omics layers for the MoTrPAC human pre-COVID suspension cohort. For each assay type, it visualizes key QC metrics and normalized expression distributions, and compares regenerated outputs to the current public data release where applicable.

Note: Reproduction of outputs from source requires access to internal MoTrPAC Google Cloud Storage buckets, which are not publicly available. Outputs and visualizations derived from public release data are accessible without GCS credentials.

---

## Setup

Load configuration, set local output directory, and source all ome-specific normalization functions.

If you're interested in rerunning and generating all the qc_norm components at once instead of calling each ome's generate function separately, see: `data-raw/generate_normalized_expression/generate_qc_norm.R`

```{r setup-paths}
config = jsonlite::fromJSON("~/config.json")
repo_local_dir = paste0(config$precovid_repo_path, "data/tmp/")

source(file.path(here(), "data-raw", "gsutil_path_parsing.R"))
source(file.path(here(), "data-raw", "generate_differential_analysis", "generate_differential_modeling_functions.R"))

invisible(lapply(
  list.files(file.path(here::here(), "data-raw", "generate_normalized_expression"),
             pattern = "\\.R$", full.names = TRUE),
  source
)
)
```

---


## Transcriptomics (RNA-seq)

RNA-seq counts are filtered (> 0.5 CPM in ≥ 10% of samples per tissue), TMM-normalized, log-CPM transformed, and batch-corrected via `limma::removeBatchEffect`. Batch-corrected matrices are for visualization only; differential analysis uses raw counts.

```{r transcriptomics-generate, eval=FALSE}
#sourced from respective file in `data-raw/generate_normalized_expression`
generate_transcriptomics_qc_norm(repo_local_dir)
```


After running, compare to existing output from `load_qc` in the `MotrpacHumanPreSuspensionData` package

```{r transcriptomics-existing}
existing_rna_package = MotrpacHumanPreSuspensionData::load_qc(selected_omes = "transcript-rna-seq",
                                                              load_acute_only = FALSE)

qc_norm_path = paste0(repo_local_dir, "freeze/transcriptomics/qc-norm/")
metadata_path = paste0(repo_local_dir, "freeze/transcriptomics/metadata/")
regenerated_rna = load_regenerated_qc("transcript-rna-seq", qc_norm_path, metadata_path)

compare_qc_norm("transcript-rna-seq", existing_rna_package, regenerated_rna)
```


---

## Proteomics

### Global Proteomics (Prot-PR)

```{r prot-pr-generate, eval=FALSE}
#sourced from respective file in `data-raw/generate_normalized_expression`
generate_prot_pr_qc_norm(repo_local_dir)
```

```{r prot-pr-visualize}
existing_prot_pr_package = MotrpacHumanPreSuspensionData::load_qc(selected_omes = "prot-pr",
                                                                  load_acute_only = FALSE)

qc_norm_path = paste0(repo_local_dir, "freeze/proteomics/qc-norm/")
metadata_path = paste0(repo_local_dir, "freeze/proteomics/metadata/")
regenerated_prot_pr = load_regenerated_qc("prot-pr", qc_norm_path, metadata_path,
                                          tissues = c("muscle", "adipose"),
                                          qc_norm_details = "log2-mn")

compare_qc_norm("prot-pr", existing_prot_pr_package, regenerated_prot_pr,
                tissues = c("muscle", "adipose"),
                feature_key = "protein_id")
```

### Phosphoproteomics (Prot-PH)

```{r prot-ph-generate, eval=FALSE}
#sourced from respective file in `data-raw/generate_normalized_expression`
generate_prot_ph_qc_norm(repo_local_dir)
```

```{r prot-ph-visualize}
existing_prot_ph_package = MotrpacHumanPreSuspensionData::load_qc(selected_omes = "prot-ph",
                                                                  load_acute_only = FALSE)

qc_norm_path = paste0(repo_local_dir, "freeze/proteomics/qc-norm/")
metadata_path = paste0(repo_local_dir, "freeze/proteomics/metadata/")
regenerated_prot_ph = load_regenerated_qc("prot-ph", qc_norm_path, metadata_path,
                                          tissues = c("muscle", "adipose"),
                                          qc_norm_details = "log2-mn")

compare_qc_norm("prot-ph", existing_prot_ph_package, regenerated_prot_ph,
                tissues = c("muscle", "adipose"),
                feature_key = "protein_id")
```

### Olink Targeted Proteomics (Prot-OL)

```{r prot-ol-generate, eval=FALSE}
#sourced from respective file in `data-raw/generate_normalized_expression`
generate_prot_ol_qc_norm(repo_local_dir)
```

```{r prot-ol-visualize}
existing_prot_ol_package = MotrpacHumanPreSuspensionData::load_qc(selected_omes = "prot-ol",
                                                                  load_acute_only = FALSE)

qc_norm_path = paste0(repo_local_dir, "freeze/proteomics/qc-norm/")
metadata_path = paste0(repo_local_dir, "freeze/proteomics/metadata/")
regenerated_prot_ol = load_regenerated_qc("prot-ol", qc_norm_path, metadata_path,
                                          tissues = "blood",
                                          qc_norm_details = "log2")

compare_qc_norm("prot-ol", existing_prot_ol_package, regenerated_prot_ol,
                tissues = "blood")
```

---

## Metabolomics

```{r metab generate, eval=FALSE}
generate_metab_qc_norm()
```

```{r metab-visualize}
existing_metab_package = MotrpacHumanPreSuspensionData::load_qc(selected_omes = "metab",
                                                                load_acute_only = FALSE,
                                                                remove_redundant_metab = FALSE)

metab_outdir = file.path(config$precovid_repo_path, "data", "tmp", "freeze")

for (tissue in names(existing_metab_package)) {
  for (platform in names(existing_metab_package[[tissue]])) {
    message(tissue); message(platform)
    subdir = ifelse(grepl("metab-t", platform), "metabolomics-targeted", "metabolomics-untargeted")
    qc_norm_path = file.path(metab_outdir, subdir, "qc-norm/")
    metadata_path = file.path(metab_outdir, subdir, "metadata/")
    
    regenerated_metab = load_regenerated_qc(platform, 
                                            qc_norm_path, 
                                            metadata_path,
                                            tissues = tissue,
                                            qc_norm_details = "log2",
                                            version = "1.4")
    
    compare_qc_norm(platform, existing_metab_package, regenerated_metab,
                    tissues = tissue,
                    feature_key = "metabolite_name")
  }
}
```

---

## Epigenomics

### ATAC-seq

No changes needed for ATACseq. feature metadata lines up perfectly. 
```{r, eval = FALSE}
atac_dataset = load_qc(selected_omes = "epigen-atac-seq",
                       epigen = TRUE,
                       repo_local_dir = repo_local_dir)
```

### Methylation (MethylCap-seq)

Skipping for file size purposes. 

---

## Session Info

```{r session-info}
sessionInfo()
```
